---
title: "simu-data-v2"
author: "JT Miller"
format: html
editor: visual
---

## Simulate Pheno Extraction and Interpolation Across a species Range

### Methods

Load Packages

```{r}
library(rnaturalearth)
library(sf)
library(ggplot2)
library(terra)
library(dplyr)
library(patchwork)
library(dbscan)
library(mgcv)
```

#### Step 1: Create Nevada Phenology Simulation with Daymet climate Data

```{r}
# get nevada boundary
nevada <- ne_states(country = "United States of America", 
                    returnclass = "sf") %>% filter(name == "Nevada")

```

```{r}
# Path to Daymet data
daymet_path <- "/orange/guralnick/daymet-daily/"

# Function to extract Nevada data from Daymet NetCDF
extract_nevada_daymet <- function(variable, years) {
  
  climate_list <- list()
  
  for(year in years) {
    # Load NetCDF file
    nc_file <- file.path(daymet_path, 
                         paste0("daymet_v4_daily_na_", variable, "_", year, ".nc"))
    climate_rast <- rast(nc_file)
    
    # Reproject Nevada to match Daymet CRS
    nevada_reproj <- st_transform(nevada, crs(climate_rast))
    nevada_vect <- vect(nevada_reproj)
    
    # Crop and mask to Nevada
    climate_nv <- crop(climate_rast, nevada_vect, mask = TRUE)
    
    climate_list[[as.character(year)]] <- climate_nv
  }
  
  # Combine all years
  climate_all <- do.call(c, climate_list)
  
  return(climate_all)
}

# Extract data for 2017-2020
years <- 2017:2020
tmin_nv <- extract_nevada_daymet("tmin", years)
tmax_nv <- extract_nevada_daymet("tmax", years)
prcp_nv <- extract_nevada_daymet("prcp", years)

# combine elements into a singular raster
tmin_nv_comined <- rast(tmin_nv)
tmax_nv_comined <- rast(tmax_nv)
prcp_nv_comined <- rast(prcp_nv)

# additionally calc tmean from tmin and tmax 
tmean_nv <- (tmin_nv_comined + tmax_nv_comined)/2

# Save for later use
writeRaster(tmin_nv_comined, "/blue/guralnick/millerjared/PhenoInterpolate/data/processed/clim-layers/nevada_tmin_2017_2020_daymet.tif", overwrite = TRUE)
writeRaster(tmax_nv_comined, "/blue/guralnick/millerjared/PhenoInterpolate/data/processed/clim-layers/nevada_tmax_2017_2020_daymet.tif", overwrite = TRUE)
writeRaster(prcp_nv_comined, "/blue/guralnick/millerjared/PhenoInterpolate/data/processed/clim-layers/nevada_prcp_2017_2020_daymet.tif", overwrite = TRUE)
writeRaster(tmean_nv, "/blue/guralnick/millerjared/PhenoInterpolate/data/processed/clim-layers/nevada_tmean_2017_2020_daymet.tif", overwrite = TRUE)

```

Read in Raster Stack

```{r}
tmin_nev <- terra::rast("/blue/guralnick/millerjared/PhenoInterpolate/data/processed/clim-layers/nevada_tmin_2017_2020_daymet.tif")
tmax_nev <- terra::rast("/blue/guralnick/millerjared/PhenoInterpolate/data/processed/clim-layers/nevada_tmin_2017_2020_daymet.tif")
tmean_nev <- terra::rast("/blue/guralnick/millerjared/PhenoInterpolate/data/processed/clim-layers/nevada_tmean_2017_2020_daymet.tif")
ppt_nev <- terra::rast("/blue/guralnick/millerjared/PhenoInterpolate/data/processed/clim-layers/nevada_prcp_2017_2020_daymet.tif")
```

Create a fake species range

```{r}
nevada_bbox <- st_bbox(nevada)

species_range <- st_polygon(list(cbind(
                        c(-120, -119, -117, -115, -116, -118, -120), # lons
                        c(36, 39, 41, 39.5, 37, 35.5, 36)))) %>% #lats
  st_sfc(crs = st_crs(nevada)) %>% 
  st_sf() %>% 
  st_intersection(nevada) # clip to nev boundary

ggplot() + 
  geom_sf(data = nevada, mapping = aes(), fill = "lightgray") + 
  geom_sf(data = species_range, mapping = aes(), fill = "darkgreen", alpha = 0.5) + 
  theme_bw() + 
  ggtitle("Species S Distribution in Nevada")
```

#### Step 2: Set-up flowering Response

```{r}
## Create a true data generating process
params <- list(
  intercept = 170,      # flowering DOY when temp = 0°C
  temp_slope = -3.5,    # days earlier per 1°C increase
  sigma_spatial = 5, # variation between sites
  sigma_individual = 2 # variation within populations
)

# Population flowering distributions
sim_pop_flowering <- function(temp, n_individuals = 100, par = params){
  # mean flowering for temp
  mu_site <- par$intercept + par$temp_slope * temp
  # generate distribution of individaul flowering times
  flowering_times <- rnorm(n_individuals, 
                           mean = mu_site, 
                           sd = par$sigma_individual)
  return(flowering_times)
}

# visualize response curve with phenology dists 
temp_range <- seq(0, 25, by = 0.5)

# generate distributions for each temp 
response_distributions <- data.frame(temp_range = temp_range) %>% 
  rowwise() %>% 
  mutate(
    # gen full flowering dist for this temp
    distribution = list(sim_pop_flowering(temp_range, 
                                          n_individuals = 500,
                                          par = params)), 
    # calc phenometrics from distribution
    onset_10 = quantile(distribution, 0.10), 
    median_50 = quantile(distribution, 0.50),
    offset_90 = quantile(distribution, 0.90)
  ) %>% ungroup() 

response_long <- response_distributions %>%
  tidyr::pivot_longer(cols = c(onset_10, median_50, offset_90),
                      names_to = "phenometric",
                      values_to = "doy") %>%
  mutate(phenometric = factor(phenometric,
                              levels = c("onset_10", "median_50", "offset_90"),
                              labels = c("Onset (10%)", "Median (50%)", "Offset (90%)")))

ggplot(response_long, aes(x = temp_range, y = doy, color = phenometric, linetype = phenometric)) +
  geom_ribbon(data = response_distributions, 
              aes(x = temp_range, ymin = onset_10, ymax = offset_90),
              fill = "yellow", alpha = 0.3, inherit.aes = FALSE) +
  geom_line(size = 1) +
  scale_color_manual(values = c("darkred", "darkorange", "steelblue")) +
  scale_linetype_manual(values = c("dashed", "solid", "dashed")) +
  labs(color = "Phenometric", linetype = "Phenometric") +
  theme_bw() +
  ggtitle("Community Response to Temperature Driven Phenology")
```

Match up species range with climate, then simu spring flowering

```{r}

# reproj species range to match clim data
species_range_clim <- st_transform(species_range, crs = crs(tmean_nev))

# create a point for each 1km cell center within the species range (res of daymet clim layer)
tmean_species <- crop(tmean_nev, vect(species_range_clim), mask = TRUE)
ppt_species <- crop(ppt_nev, vect(species_range_clim), mask = TRUE)
# extract cell centers as pts 
cells <- cells(tmean_species[[1]]) # get cell numbers
cell_coords <- xyFromCell(tmean_species[[1]], cells)
cell_pts <- st_as_sf(data.frame(cell_coords), 
                    coords = c("x", "y"), 
                    crs = crs(tmean_nev))

# calc spring clim at each cell
spring_layers <- 30:150 # Feb-May
tmean_spring <- mean(tmean_species[[spring_layers]])
ppt_spring <- sum(ppt_species[[spring_layers]])

# extract clim vals at each cell
climate_data <- terra::extract(c(tmean_spring, ppt_spring), 
                               vect(cell_pts))

# calc TRUE flowering distributions for each cell
cells_true_flowering__dists <- cell_pts %>% 
  st_drop_geometry() %>% 
  bind_cols(st_coordinates(cell_pts)) %>% 
  mutate(
    tmean_spring = climate_data$mean,
    ppt_spring = climate_data$sum, 
    year = 2017
  ) %>% 
  rowwise() %>% 
  mutate(
    cell_flowering_distribution = list(sim_pop_flowering(tmean_spring, n_individuals = 100, par = params)
  ), 
  cell_flowering_onset = quantile(cell_flowering_distribution, 0.10), 
  cell_flowering_median = quantile(cell_flowering_distribution, 0.50), 
  cell_flowering_offset = quantile(cell_flowering_distribution, 0.90)) %>% 
  ungroup() 
  

# vis against response
onset_response_long <- response_long %>% filter(phenometric == "Onset (10%)")

ggplot() +
  geom_point(cells_true_flowering__dists, mapping = aes(x = tmean_spring, y = cell_flowering_onset)) +
  geom_line(onset_response_long, mapping = aes(x = temp_range, y = doy), size = 1.2, color = "darkred", alpha = 0.1) +
  labs(x = "Mean Spring Temperature (°C)", 
       y = "Flowering Onset (DOY)",
       title = "True Flowering Response to Temperature") +
  theme_bw() +
  theme(text = element_text(size = 14))

ggsave("/blue/guralnick/millerjared/PhenoInterpolate/outputs/figures/true-fl-response-curve.png")


# create a spatial plot: 
cells_true_flowering_dists_sf <- st_as_sf(cells_true_flowering__dists, 
                                coords = c("X", "Y"), 
                                crs = crs(tmean_nev))

true_flowering_plot <- ggplot() + 
  geom_sf(data = nevada, fill = "gray90", color = "black") + 
  geom_sf(data = species_range, fill = NA, color = "black", size = 1) + 
  geom_sf(data = cells_true_flowering_dists_sf, mapping = aes(color = cell_flowering_onset), size = 0.5) + 
  scale_color_viridis_c(option = "magma", direction = -1, name = "Flowering\nDOY") + 
  ggtitle("True Flowering Onset (10%) Across Species Range")

temp_plot <- ggplot() + 
  geom_sf(data = nevada, fill = "gray90", color = "black") + 
  geom_sf(data = species_range, fill = NA, color = "black", size = 1) + 
  geom_sf(data = cells_true_flowering_dists_sf, mapping = aes(color = tmean_spring), size = 0.5) + 
  scale_color_viridis_c(option = "plasma", name = "Mean Spring Temp") + 
  ggtitle("Mean Spring Temperature Across Species Range")

combined_plot <- true_flowering_plot + temp_plot

ggsave(plot = combined_plot, "/blue/guralnick/millerjared/PhenoInterpolate/outputs/figures/nev-sp-simu-map-plots.png", width = 15, height = 10)
print(combined_plot)


## Do this for each year 
years <- 2017:2020
days_per_year <- c(365,365,365,366)
spring_doy <- 30:150

# calc cumulative yera positions 
year_starts <- c(1, cumsum(days_per_year[-length(days_per_year)]) + 1)
names(year_starts) <- years

# genearte true occurrences for all years
cells_true_flowering_all <- list()
for(i in seq_along(years)){
  # get year and when year starts in raster
  year <- years[i]
  year_start <- year_starts[i]
  # get spring layers for this year
  spring_layers_year <- (year_start - 1) + spring_doy
  # calc spring climate for this year
  tmean_spring <- mean(tmean_species[[spring_layers_year]])
  ppt_spring <- sum(ppt_species[[spring_layers_year]])
  # extract clim vals 
  climate_data <- terra::extract(c(tmean_spring, ppt_spring), vect(cell_pts))
  # generate cell phenometrics per year
  cells_true_flowering__dists_year <- cell_pts %>% 
  st_drop_geometry() %>% 
  bind_cols(st_coordinates(cell_pts)) %>% 
  mutate(
    tmean_spring = climate_data$mean,
    ppt_spring = climate_data$sum, 
    year = year
  ) %>% 
  rowwise() %>% 
  mutate(
    cell_flowering_distribution = list(sim_pop_flowering(tmean_spring, n_individuals = 100, par = params)
  ), 
  cell_flowering_onset = quantile(cell_flowering_distribution, 0.10), 
  cell_flowering_median = quantile(cell_flowering_distribution, 0.50), 
  cell_flowering_offset = quantile(cell_flowering_distribution, 0.90)) %>% 
  ungroup() 
  cells_true_flowering_all[[as.character(year)]] <- cells_true_flowering__dists_year
 
}

# combine all years
cells_true_flowering <- bind_rows(cells_true_flowering_all)
cells_true_flowering_sf  <- st_as_sf(cells_true_flowering, 
                                coords = c("X", "Y"), 
                                crs = crs(tmean_nev))


true_flowering_plot_all <- ggplot() + 
  geom_sf(data = nevada, fill = "gray90", color = "black") + 
  geom_sf(data = species_range, fill = NA, color = "black", size = 1) + 
  geom_sf(data = cells_true_flowering_sf , mapping = aes(color = cell_flowering_onset), size = 0.5) + 
  scale_color_viridis_c(option = "magma", direction = -1, name = "Flowering\nDOY") + 
  facet_wrap(~year) +
  ggtitle("True Flowering Onset (10%) Across Species Range") +
  theme_bw()
  
print(true_flowering_plot_all)

```

#### Step 3: Create a sampling layer

```{r}
# start simple, lets evenly sample cells across space and across flowering distributions
n_cells_per_year <- 50
obs_per_cell_range <- c(5,20)

# sample from true pops 
observed_occs <- cells_true_flowering %>% 
  group_by(year) %>% 
  # randomly sample cells 
  slice_sample(n = n_cells_per_year, replace = FALSE) %>% 
  ungroup() %>% 
  rowwise() %>% 
  mutate( # randomly determine how many obs within cell
    n_obs_at_cell = sample(obs_per_cell_range[1]:obs_per_cell_range[2], 1), 
    # sample flowering from distribution
    observed_flowering_times = list(sample(cell_flowering_distribution, n_obs_at_cell))
  ) %>% 
  ungroup() %>% 
  tidyr::unnest(observed_flowering_times) %>%  # each row will be an obs
  rename(flowering_doy_obs = observed_flowering_times)

# visualize sampling 
# create a spatial plot: 
observed_occs_sf <- st_as_sf(observed_occs, 
                                coords = c("X", "Y"), 
                                crs = crs(tmean_nev))

observed_occs_sf <- observed_occs_sf %>% 
  distinct(geometry, year, n_obs_at_cell)

obs_flowering_plot <- ggplot() + 
  geom_sf(data = nevada, fill = "gray90", color = "black") + 
  geom_sf(data = species_range, fill = NA, color = "black", size = 1) + 
  geom_sf(data = observed_occs_sf, mapping = aes(color = n_obs_at_cell), size = 0.5) + 
  scale_color_viridis_c(option = "magma", direction = +1, name = "Number of Obs") + 
  ggtitle("Flowering Observation Simu") + 
  facet_wrap(~year)

obs_flowering_plot 

```

#### Step 4: Group clusters to build phenometric estimates

```{r}
## Spatial Clustering
cluster_spatial <- function(data, distance_km = 50, min_obs = 7, min_cells = 3){
  data_clustered <- data %>% 
    group_by(year) %>% 
    group_modify(~{
      # get unique cell locs for clustering
      unique_cells <- .x %>% 
        distinct(X,Y, .keep_all = TRUE) %>% 
        dplyr::select(X,Y, tmean_spring, ppt_spring)
      coords <- as.matrix(unique_cells[, c("X", "Y")])
      # convert dist to meters
      eps_m <- distance_km * 1000
      # DBSCAN clustering on cells, instead of individual obs
      clusters <- dbscan(coords, eps = eps_m, minPts = min_cells)
      # add cluster IDs to unique cells
      unique_cells$cluster_id <- clusters$cluster
      # join cluster IDs back to all obs
      .x %>% 
        left_join(unique_cells %>% dplyr::select(X,Y,cluster_id), by = c("X","Y")) %>% 
        filter(cluster_id != 0) # removes noise cluster
    }) %>% 
    ungroup() %>% 
    # create unique cluster id across years
    mutate(cluster_full_id = paste0(year, "_", cluster_id)) %>% 
    # filter to clusters with min total obs
    group_by(cluster_full_id) %>% 
    filter(n() >= min_obs) %>% 
    # calc spatial metrics per cluster
    mutate(
      cluster_n_obs = n(), # total obs in cluster
      cluster_n_cells = n_distinct(X,Y), # number of cells in cluster
      # Calc cell level centroids for spatial metrics 
      cluster_mean_dist_to_center = {
        cell_coords <- distinct(pick(X,Y))
        centroid_x <- mean(cell_coords$X)
        centroid_y <- mean(cell_coords$Y)
        sqrt(mean((cell_coords$X - centroid_x)^2 + (cell_coords$Y - centroid_y)^2)) / 1000
      }, 
      # mean nearest neighbor dist between cells (km)
      cluster_mean_nn_dist = {
        cell_coords <- distinct(pick(X, Y))
        if(nrow(cell_coords) > 1) {
          coords_mat <- as.matrix(cell_coords)
          dist_mat <- as.matrix(dist(coords_mat))
          diag(dist_mat) <- NA
          mean(apply(dist_mat, 1, min, na.rm = TRUE)) / 1000
        } else {
          NA_real_
        }
      },
      
      # Inverse for weighting (lower distance = higher weight)
      cluster_compactness = 1 / (cluster_mean_dist_to_center + 0.1)
    ) %>%
    ungroup()
  
  return(data_clustered)
}


## Environmental Clustering 
cluster_env <- function(data, temp_range_c = 2, min_obs = 7, min_cells = 3){
  data_clustered <- data %>% 
    group_by(year) %>% 
    group_modify(~{
      # get unique cell locations with their temps 
      unique_cells <- .x %>% 
        distinct(X,Y, .keep_all = TRUE) %>% 
        dplyr::select(X,Y,tmean_spring, ppt_spring)
      # run hierarchical clustering on cell temp
      temp_dist <- dist(unique_cells$tmean_spring)
      hc <- hclust(temp_dist, method = "complete")
      # cut tree to get clusters with max temp_range_c difference
      clusters <- cutree(hc, h = temp_range_c)
      # add cluster IDs to unique cells
      unique_cells$cluster_id <- clusters
      # join cliuster IDs back to all obs
      .x %>% 
        left_join(unique_cells %>% dplyr::select(X,Y,cluster_id), by = c("X", "Y"))
        
    }) %>% 
    ungroup() %>% 
    # create unique cluster IDs across years
    mutate(cluster_full_id = paste0(year, "_", cluster_id)) %>% 
    # Filter to clusters with min obs
    group_by(cluster_full_id) %>% 
    filter(n_distinct(X,Y) >= min_cells, 
           n() >= min_obs) %>% 
    # Calculate environmental variation per cluster at cell  level
   mutate(
      cluster_n_obs = n(),  # Total observations in cluster
      cluster_n_cells = n_distinct(X, Y),  # Number of cells in cluster
      
      # Temperature metrics calculated from unique cells
      cluster_temp_mean = {
        cell_temps <- distinct(pick(X, Y, tmean_spring))$tmean_spring
        mean(cell_temps)
      },
      cluster_temp_sd = {
        cell_temps <- distinct(pick(X, Y, tmean_spring))$tmean_spring
        sd(cell_temps)
      },
      cluster_temp_range = {
        cell_temps <- distinct(pick(X, Y, tmean_spring))$tmean_spring
        max(cell_temps) - min(cell_temps)
      },
      cluster_temp_cv = {
        cell_temps <- distinct(pick(X, Y, tmean_spring))$tmean_spring
        sd(cell_temps) / abs(mean(cell_temps))
      },
      
      # Weight: prefer environmentally homogeneous clusters
      cluster_env_weight = 1 / (cluster_temp_sd + 0.1)
    ) %>%
    ungroup()
  
  return(data_clustered)
}
obs_spatial_clusters <- cluster_spatial(observed_occs, 
                                        distance_km = 25, 
                                        min_obs = 7,     
                                        min_cells = 2)    
obs_env_clusters <- cluster_env(observed_occs, 
                                temp_range_c = 2, 
                                min_obs = 7, 
                                min_cells = 2)

# Visualize clusters in space
obs_spatial_clusters_sf <- st_as_sf(obs_spatial_clusters, 
                                     coords = c("X", "Y"), 
                                     crs = crs(tmean_nev))
obs_env_clusters_sf <- st_as_sf(obs_env_clusters, 
                                     coords = c("X", "Y"), 
                                     crs = crs(tmean_nev))

ggplot() +
  geom_sf(data = nevada, fill = "gray90") +
  geom_sf(data = obs_spatial_clusters_sf, aes(color = factor(cluster_id)), size = 1) +
  facet_wrap(~year) +
  labs(color = "Cluster", title = "Spatial Clusters by Year") +
  theme_minimal()

ggplot() +
  geom_sf(data = nevada, fill = "gray90") +
  geom_sf(data = obs_env_clusters_sf, aes(color = factor(cluster_id)), size = 1) +
  facet_wrap(~year) +
  labs(color = "Cluster", title = "Temp Env Clusters by Year") +
  theme_minimal()
```

#### Step 5: Use phenesse to Calc Onset

```{r}
cluster_spatial_onset <- obs_spatial_clusters %>%
  group_by(cluster_full_id, year, cluster_compactness) %>%
  summarise(
    n_obs = n(),
    onset_estimate = phenesse::quantile_ci(observations = flowering_doy_obs, 
                                  percentile = 0.10, 
                                  bootstraps = 250)$estimate,
    # Also get cluster-level environmental predictors for GAM
    tmean_spring_mean = mean(tmean_spring),
    ppt_spring_mean = mean(ppt_spring),
    X_centroid = mean(X),
    Y_centroid = mean(Y),
    .groups = "drop"
  )  %>% 
  mutate(weight = cluster_compactness / mean(cluster_compactness)) # normalized

cluster_env_onset <- obs_env_clusters %>%
  group_by(cluster_full_id, year, cluster_temp_sd) %>%
  summarise(
    n_obs = n(),
    onset_estimate = phenesse::quantile_ci(observations = flowering_doy_obs, 
                                  percentile = 0.10, 
                                  bootstraps = 250)$estimate,
    # Also get cluster-level environmental predictors for GAM
    tmean_spring_mean = mean(tmean_spring),
    ppt_spring_mean = mean(ppt_spring),
    X_centroid = mean(X),
    Y_centroid = mean(Y),
    .groups = "drop"
  )  %>% 
  mutate(weight = mean(cluster_temp_sd)/cluster_temp_sd)

```

#### Step 6. Model Onset using GAM

```{r}
cluster_spatial_onset$cluster_full_id <- as.factor(cluster_spatial_onset$cluster_full_id)
cluster_env_onset$cluster_full_id <- as.factor(cluster_env_onset$cluster_full_id)
# run GAM, overkill here.
gam_spatial <- gam(onset_estimate ~ s(tmean_spring_mean, k = 5) + 
                     s(cluster_full_id, bs = "re"), 
                   data = cluster_spatial_onset, 
                   weights = weight)

gam_env <- gam(onset_estimate ~ s(tmean_spring_mean, k = 5) + 
                     s(cluster_full_id, bs = "re"), 
                   data = cluster_env_onset, 
                   weights = weight)

# predict 
cells_true_flowering$predicted_onset_spatial <- predict(gam_spatial, newdata = data.frame(
                        tmean_spring_mean = cells_true_flowering$tmean_spring), 
                        exclude = "s(cluster_full_id)",  # Exclude random effect
                        newdata.guaranteed = TRUE)

cells_true_flowering$predicted_onset_env <- predict(gam_env, newdata = data.frame(
                        tmean_spring_mean = cells_true_flowering$tmean_spring), 
                        exclude = "s(cluster_full_id)",  # Exclude random effect
                        newdata.guaranteed = TRUE)


```

#### Step 7. Calculate MESS metrics

```{r}
# predictor vars to check 
predictor_vars <- c("tmean_spring") # ill add ppt_spring later

# set-up training data (empirical data) for the models
training_cells_spatial <- obs_spatial_clusters %>% 
  distinct(X, Y, .keep_all = TRUE)
training_cells_env <- obs_env_clusters %>% 
  distinct(X, Y, .keep_all = TRUE)

# Spatial Model MESS
for(v in 1:length(predictor_vars)){
  # pull out var
  var_name <- predictor_vars[v]
  # define boundary extents of training data
  low  <- min(training_cells_spatial[[var_name]], na.rm = TRUE)
  high <- max(training_cells_spatial[[var_name]], na.rm = TRUE)
  # binary flags for interpolate (1) vs extrapolate (0)
  cells_true_flowering[[paste0("mess_spatial_", var_name)]] <-
    ifelse(cells_true_flowering[[var_name]] < low | cells_true_flowering[[var_name]] > high, 0, 1)
  # calculate the distance to the threshold boundaries
  cells_true_flowering[[paste0("mess_dist_spatial_", var_name)]] <-
    pmin(abs(cells_true_flowering[[var_name]] - low),
         abs(cells_true_flowering[[var_name]] - high))
  # set dist to 0 if within range (interpolation)
  cells_true_flowering[[paste0("mess_dist_spatial_", var_name)]] <-
    ifelse(cells_true_flowering[[paste0("mess_spatial_", var_name)]] == 1,
           0, cells_true_flowering[[paste0("mess_dist_spatial_", var_name)]])
}

mess_sf <- st_as_sf(cells_true_flowering, coords = c("X","Y"), crs = crs(tmean_nev))

mess_plot <- ggplot() + 
  geom_sf(data = nevada, fill = "gray90", color = "black") + 
  geom_sf(data = species_range, fill = NA, color = "black", size = 1) + 
  geom_sf(data = mess_sf , mapping = aes(color = mess_dist_spatial_tmean_spring), size = 0.5) + 
  scale_color_gradientn(
    colors = c("darkgreen", "yellow", "darkred"),
    values = scales::rescale(c(0, 0.001, max(mess_sf$mess_dist_spatial_tmean_spring, na.rm = TRUE))),
    name = "Extrapolation\nDistance"
  ) +
  facet_wrap(~year) +
  ggtitle("MESS Map: Extrapolation of tmean_spring with Spatial Clustering") +
  theme_bw()

print(mess_plot)

ggsave(plot = mess_plot, file = "/blue/guralnick/millerjared/PhenoInterpolate/outputs/figures/MESS-spatial-simu.png", width = 15, height = 15)
for(v in 1:length(predictor_vars)){
  # pull out var
  var_name <- predictor_vars[v]
  # define boundary extents of training data
  low  <- min(training_cells_env[[var_name]], na.rm = TRUE)
  high <- max(training_cells_env[[var_name]], na.rm = TRUE)
  # binary flags for interpolate (1) vs extrapolate (0)
  cells_true_flowering[[paste0("mess_env_", var_name)]] <-
    ifelse(cells_true_flowering[[var_name]] < low | cells_true_flowering[[var_name]] > high, 0, 1)
  # calculate the distance to the threshold boundaries
  cells_true_flowering[[paste0("mess_dist_env_", var_name)]] <-
    pmin(abs(cells_true_flowering[[var_name]] - low),
         abs(cells_true_flowering[[var_name]] - high))
  # set dist to 0 if within range (interpolation)
  cells_true_flowering[[paste0("mess_dist_env_", var_name)]] <-
    ifelse(cells_true_flowering[[paste0("mess_env_", var_name)]] == 1,
           0, cells_true_flowering[[paste0("mess_dist_env_", var_name)]])
}

mess_sf <- st_as_sf(cells_true_flowering, coords = c("X","Y"), crs = crs(tmean_nev))

mess_plot <- ggplot() + 
  geom_sf(data = nevada, fill = "gray90", color = "black") + 
  geom_sf(data = species_range, fill = NA, color = "black", size = 1) + 
  geom_sf(data = mess_sf , mapping = aes(color = mess_dist_env_tmean_spring), size = 0.5) + 
  scale_color_gradientn(
    colors = c("darkgreen", "yellow", "darkred"),
    values = scales::rescale(c(0, 0.001, max(mess_sf$mess_dist_env_tmean_spring, na.rm = TRUE))),
    name = "Extrapolation\nDistance"
  ) +
  facet_wrap(~year) +
  ggtitle("MESS Map: Extrapolation of tmean_spring with Enviromental Clustering") +
  theme_bw()

print(mess_plot)


ggsave(plot = mess_plot, file = "/blue/guralnick/millerjared/PhenoInterpolate/outputs/figures/MESS-env -simu.png", width = 15, height = 15)

```

#### Step 8. Calc Preformance Metrics

```{r}
#### Step 8: Error Analysis by Interpolation/Extrapolation Status

# --- Calculate absolute error for each model ---
cells_true_flowering <- cells_true_flowering %>%
  mutate(
    error_spatial = predicted_onset_spatial - cell_flowering_onset,
    error_env     = predicted_onset_env     - cell_flowering_onset,
    abs_error_spatial = abs(error_spatial),
    abs_error_env     = abs(error_env),
    interp_status_spatial = ifelse(mess_spatial_tmean_spring == 1, "Interpolation", "Extrapolation"),
    interp_status_env     = ifelse(mess_env_tmean_spring     == 1, "Interpolation", "Extrapolation")
  )

# --- Plot 1: Absolute error by interpolation status (boxplot) ---
error_long <- cells_true_flowering %>%
  dplyr::select(year, abs_error_spatial, abs_error_env,
                interp_status_spatial, interp_status_env) %>%
  tidyr::pivot_longer(
    cols = c(abs_error_spatial, abs_error_env),
    names_to = "model",
    values_to = "abs_error"
  ) %>%
  mutate(
    interp_status = ifelse(model == "abs_error_spatial", interp_status_spatial, interp_status_env),
    model = recode(model,
                   "abs_error_spatial" = "Spatial Clustering",
                   "abs_error_env"     = "Environmental Clustering")
  )

error_box_plot <- ggplot(error_long, aes(x = interp_status, y = abs_error, fill = interp_status)) +
  geom_boxplot(outlier.alpha = 0.3) +
  scale_fill_manual(values = c("Interpolation" = "darkgreen", "Extrapolation" = "darkred")) +
  facet_grid(model ~ year) +
  labs(x = NULL, y = "Absolute Error (DOY)",
       title = "Prediction Error: Interpolation vs. Extrapolation",
       fill = "Status") +
  theme_bw()
print(error_box_plot)
ggsave(plot = error_box_plot,
       file = "/blue/guralnick/millerjared/PhenoInterpolate/outputs/figures/error-by-interp-status.png",
       width = 15, height = 8)

# --- Plot 2: Response curves - predicted vs true, colored by interp status ---
# build true response curve for overlay
true_curve <- data.frame(tmean_spring = seq(
  min(cells_true_flowering$tmean_spring, na.rm = TRUE),
  max(cells_true_flowering$tmean_spring, na.rm = TRUE),
  length.out = 200
)) %>%
  mutate(true_onset = qnorm(0.10, # we're looking at the 10th percentile for onset
                            mean = params$intercept + params$temp_slope * tmean_spring,
                            sd = params$sigma_individual))

response_curve_spatial <- ggplot(cells_true_flowering,
                                  aes(x = tmean_spring, y = predicted_onset_spatial,
                                      color = interp_status_spatial)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_line(data = true_curve, aes(x = tmean_spring, y = true_onset),
            color = "black", linewidth = 1.2, inherit.aes = FALSE, alpha = 0.5) +
  scale_color_manual(values = c("Interpolation" = "darkgreen", "Extrapolation" = "darkred")) +
  facet_wrap(~year) +
  labs(x = "Mean Spring Temperature (°C)", y = "Flowering Onset (DOY)",
       title = "Spatial Model: Predicted vs True Response Curve",
       color = "Status") +
  theme_bw()

response_curve_env <- ggplot(cells_true_flowering,
                              aes(x = tmean_spring, y = predicted_onset_env,
                                  color = interp_status_env)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_line(data = true_curve, aes(x = tmean_spring, y = true_onset),
            color = "black", linewidth = 1.2, inherit.aes = FALSE, alpha = 0.5) +
  scale_color_manual(values = c("Interpolation" = "darkgreen", "Extrapolation" = "darkred")) +
  facet_wrap(~year) +
  labs(x = "Mean Spring Temperature (°C)", y = "Flowering Onset (DOY)",
       title = "Environmental Model: Predicted vs True Response Curve",
       color = "Status") +
  theme_bw()

combined_response <- response_curve_spatial / response_curve_env
ggsave(plot = combined_response,
       file = "/blue/guralnick/millerjared/PhenoInterpolate/outputs/figures/response-curves-interp-status.png",
       width = 15, height = 12)
print(combined_response)

# --- Summary table: mean absolute error by status and model ---
error_summary <- error_long %>%
  group_by(model, interp_status, year) %>%
  summarise(
    mean_abs_error = round(mean(abs_error, na.rm = TRUE), 2),
    median_abs_error = round(median(abs_error, na.rm = TRUE), 2),
    n_cells = n(),
    .groups = "drop"
  )
print(error_summary)
```


Compare GAM to true model fit 
```{r}

# --- Build prediction grid over observed predictor ranges ---
pred_grid <- expand.grid(
  tmean_spring_mean = seq(min(cells_true_flowering$tmean_spring, na.rm = TRUE),
                          max(cells_true_flowering$tmean_spring, na.rm = TRUE),
                          length.out = 100)
)

# --- Get GAM predictions with SE ---
pred_grid$predicted_spatial <- predict(gam_spatial,
                                        newdata = pred_grid,
                                        exclude = "s(cluster_full_id)",
                                        newdata.guaranteed = TRUE,
                                        se.fit = FALSE)
pred_grid$predicted_env <- predict(gam_env,
                                    newdata = pred_grid,
                                    exclude = "s(cluster_full_id)",
                                    newdata.guaranteed = TRUE,
                                    se.fit = FALSE)

# --- Build true curve for same grid ---
pred_grid <- pred_grid %>%
  rowwise() %>%
  mutate(
   
    true_onset = qnorm(0.10,
                       mean = params$intercept + params$temp_slope * tmean_spring_mean,
                       sd = params$sigma_individual)
  ) %>%
  ungroup()

# --- Pivot to long for faceted plotting ---
pred_long <- pred_grid %>%
  tidyr::pivot_longer(cols = c(predicted_spatial, predicted_env, true_onset),
               names_to = "source",
               values_to = "onset") %>%
  mutate(source = recode(source,
                         "predicted_spatial" = "GAM (Spatial)",
                         "predicted_env"     = "GAM (Environmental)",
                         "true_onset"        = "True Response"))

# --- Plot ---
gam_vs_true <- ggplot(pred_long, aes(x = tmean_spring_mean, y = onset,
                                      color = source, linetype = source)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("GAM (Spatial)"       = "steelblue",
                                 "GAM (Environmental)" = "darkorange",
                                 "True Response"            = "black")) +
  scale_linetype_manual(values = c("GAM (Spatial)"       = "dashed",
                                    "GAM (Environmental)" = "dashed",
                                    "True Response"            = "solid")) +
  labs(x = "Mean Spring Temperature (°C)", y = "Flowering Onset (DOY)",
       title = "GAM Estimated vs True Response Curve",
       color = "Source", linetype = "Source") +
  theme_bw()

print(gam_vs_true)
```


