---
title: "simu-data"
author: "JT Miller"
format: html
editor: visual
---

## Simulate Pheno Extraction and Interpolation Across a species Range

### Methods

Load Packages

```{r}
library(rnaturalearth)
library(sf)
library(ggplot2)
library(terra)
library(dplyr)
library(patchwork)
library(dbscan)
library(mgcv)
```

#### Step 1: Create Nevada Phenology Simulation with Daymet climate Data

```{r}
# get nevada boundary
nevada <- ne_states(country = "United States of America", 
                    returnclass = "sf") %>% filter(name == "Nevada")

```

```{r}
# Path to Daymet data
daymet_path <- "/orange/guralnick/daymet-daily/"

# Function to extract Nevada data from Daymet NetCDF
extract_nevada_daymet <- function(variable, years) {
  
  climate_list <- list()
  
  for(year in years) {
    # Load NetCDF file
    nc_file <- file.path(daymet_path, 
                         paste0("daymet_v4_daily_na_", variable, "_", year, ".nc"))
    climate_rast <- rast(nc_file)
    
    # Reproject Nevada to match Daymet CRS
    nevada_reproj <- st_transform(nevada, crs(climate_rast))
    nevada_vect <- vect(nevada_reproj)
    
    # Crop and mask to Nevada
    climate_nv <- crop(climate_rast, nevada_vect, mask = TRUE)
    
    climate_list[[as.character(year)]] <- climate_nv
  }
  
  # Combine all years
  climate_all <- do.call(c, climate_list)
  
  return(climate_all)
}

# Extract data for 2017-2020
years <- 2017:2020
tmin_nv <- extract_nevada_daymet("tmin", years)
tmax_nv <- extract_nevada_daymet("tmax", years)
prcp_nv <- extract_nevada_daymet("prcp", years)

# combine elements into a singular raster
tmin_nv_comined <- rast(tmin_nv)
tmax_nv_comined <- rast(tmax_nv)
prcp_nv_comined <- rast(prcp_nv)

# additionally calc tmean from tmin and tmax 
tmean_nv <- (tmin_nv_comined + tmax_nv_comined)/2

# Save for later use
writeRaster(tmin_nv_comined, "/blue/guralnick/millerjared/PhenoInterpolate/data/processed/clim-layers/nevada_tmin_2017_2020_daymet.tif", overwrite = TRUE)
writeRaster(tmax_nv_comined, "/blue/guralnick/millerjared/PhenoInterpolate/data/processed/clim-layers/nevada_tmax_2017_2020_daymet.tif", overwrite = TRUE)
writeRaster(prcp_nv_comined, "/blue/guralnick/millerjared/PhenoInterpolate/data/processed/clim-layers/nevada_prcp_2017_2020_daymet.tif", overwrite = TRUE)
writeRaster(tmean_nv, "/blue/guralnick/millerjared/PhenoInterpolate/data/processed/clim-layers/nevada_tmean_2017_2020_daymet.tif", overwrite = TRUE)

```

Read in Raster Stack

```{r}
tmin_nev <- terra::rast("/blue/guralnick/millerjared/PhenoInterpolate/data/processed/clim-layers/nevada_tmin_2017_2020_daymet.tif")
tmax_nev <- terra::rast("/blue/guralnick/millerjared/PhenoInterpolate/data/processed/clim-layers/nevada_tmin_2017_2020_daymet.tif")
tmean_nev <- terra::rast("/blue/guralnick/millerjared/PhenoInterpolate/data/processed/clim-layers/nevada_tmean_2017_2020_daymet.tif")
ppt_nev <- terra::rast("/blue/guralnick/millerjared/PhenoInterpolate/data/processed/clim-layers/nevada_prcp_2017_2020_daymet.tif")
```

Create a fake species range

```{r}
nevada_bbox <- st_bbox(nevada)

species_range <- st_polygon(list(cbind(
                        c(-120, -119, -117, -115, -116, -118, -120), # lons
                        c(36, 39, 41, 39.5, 37, 35.5, 36)))) %>% #lats
  st_sfc(crs = st_crs(nevada)) %>% 
  st_sf() %>% 
  st_intersection(nevada) # clip to nev boundary

ggplot() + 
  geom_sf(data = nevada, mapping = aes(), fill = "lightgray") + 
  geom_sf(data = species_range, mapping = aes(), fill = "darkgreen", alpha = 0.5) + 
  theme_bw() + 
  ggtitle("Species S Distribution in Nevada")
```

#### Step 2: Set-up flowering Response

```{r}
## Create a true data generating process
params <- list(
  intercept = 170,      # flowering DOY when temp = 0°C
  temp_slope = -3.5,    # days earlier per 1°C increase
  sigma = 0            # within-site SD
)

# True flowering DOY function (linear only)
true_flowering_mu <- function(temp, par = params){
  mu <- par$intercept + par$temp_slope * temp
  return(mu)
}

# Visualize the response curve
temp_range <- seq(0, 25, by = 0.1)
flowering_doy <- true_flowering_mu(temp_range)
response_df <- data.frame(temp_range, flowering_doy)

ggplot(response_df, mapping = aes(x = temp_range, y = flowering_doy)) +
  geom_line(size = 1.2, color = "darkred") +
  labs(x = "Mean Spring Temperature (°C)", 
       y = "Flowering Onset (DOY)",
       title = "True Flowering Response to Temperature") +
  theme_bw() +
  theme(text = element_text(size = 14))

```

Match up species range with climate, then simu spring flowering

```{r}

# reproj species range to match clim data
species_range_clim <- st_transform(species_range, crs = crs(tmean_nev))

# create a point for each 1km cell center within the species range (res of daymet clim layer)
tmean_species <- crop(tmean_nev, vect(species_range_clim), mask = TRUE)
ppt_species <- crop(ppt_nev, vect(species_range_clim), mask = TRUE)
# extract cell centers as pts 
cells <- cells(tmean_species[[1]]) # get cell numbers
cell_coords <- xyFromCell(tmean_species[[1]], cells)
occ_pts <- st_as_sf(data.frame(cell_coords), 
                    coords = c("x", "y"), 
                    crs = crs(tmean_nev))

# calc spring clim at each point
spring_layers <- 30:150 # Feb-May
tmean_spring <- mean(tmean_species[[spring_layers]])
ppt_spring <- sum(ppt_species[[spring_layers]])

# extract clim vals at each occ pt 
climate_data <- terra::extract(c(tmean_spring, ppt_spring), 
                               vect(occ_pts))

# calc TRUE flowering DOY for simulation occurrences
true_occurrences <- occ_pts %>% 
  st_drop_geometry() %>% 
  bind_cols(st_coordinates(occ_pts)) %>% 
  mutate(
    tmean_spring = climate_data$mean,
    ppt_spring = climate_data$sum, 
    flowering_doy_mu = true_flowering_mu(tmean_spring, par = params), # TRUE flowering by expected rules
    flowering_doy = rnorm(n(), flowering_doy_mu, params$sigma), # TRUE flowering by allowing SD
    year = 2017
  )

# vis against response

ggplot() +
  geom_point(true_occurrences, mapping = aes(x = tmean_spring, y = flowering_doy)) +
  geom_line(response_df, mapping = aes(x = temp_range, y = flowering_doy), size = 1.2, color = "darkred", alpha = 0.1) +
  labs(x = "Mean Spring Temperature (°C)", 
       y = "Flowering Onset (DOY)",
       title = "True Flowering Response to Temperature") +
  theme_bw() +
  theme(text = element_text(size = 14))

ggsave("/blue/guralnick/millerjared/PhenoInterpolate/outputs/figures/true-fl-response-curve.png")


# create a spatial plot: 
true_occurrences_sf <- st_as_sf(true_occurrences, 
                                coords = c("X", "Y"), 
                                crs = crs(tmean_nev))

true_flowering_plot <- ggplot() + 
  geom_sf(data = nevada, fill = "gray90", color = "black") + 
  geom_sf(data = species_range, fill = NA, color = "black", size = 1) + 
  geom_sf(data = true_occurrences_sf, mapping = aes(color = flowering_doy), size = 0.5) + 
  scale_color_viridis_c(option = "magma", direction = -1, name = "Flowering\nDOY") + 
  ggtitle("True Flowering DOY Across Species Range")

temp_plot <- ggplot() + 
  geom_sf(data = nevada, fill = "gray90", color = "black") + 
  geom_sf(data = species_range, fill = NA, color = "black", size = 1) + 
  geom_sf(data = true_occurrences_sf, mapping = aes(color = tmean_spring), size = 0.5) + 
  scale_color_viridis_c(option = "plasma", name = "Mean Spring Temp") + 
  ggtitle("Mean Spring Temperature Across Species Range")

combined_plot <- true_flowering_plot + temp_plot

ggsave(plot = combined_plot, "/blue/guralnick/millerjared/PhenoInterpolate/outputs/figures/nev-sp-simu-map-plots.png", width = 15, height = 10)
print(combined_plot)


## Do this for each year 
years <- 2017:2020
days_per_year <- c(365,365,365,366)
spring_doy <- 30:150

# calc cumulative yera positions 
year_starts <- c(1, cumsum(days_per_year[-length(days_per_year)]) + 1)
names(year_starts) <- years

# genearte true occurrences for all years
true_occurrences_all <- list()
for(i in seq_along(years)){
  # get year and when year starts in raster
  year <- years[i]
  year_start <- year_starts[i]
  # get spring layers for this year
  spring_layers_year <- (year_start - 1) + spring_doy
  # calc spring climate for this year
  tmean_spring <- mean(tmean_species[[spring_layers_year]])
  ppt_spring <- sum(ppt_species[[spring_layers_year]])
  # extract clim vals 
  climate_data <- terra::extract(c(tmean_spring, ppt_spring), vect(occ_pts))
  # generate occs for this year
  true_occurrences_year <- occ_pts %>% 
    st_drop_geometry() %>% 
    bind_cols(st_coordinates(occ_pts)) %>% 
    mutate(
      tmean_spring = climate_data$mean, 
      ppt_spring = climate_data$sum, 
       flowering_doy_mu = true_flowering_mu(tmean_spring, par = params), # TRUE flowering by expected rules
    flowering_doy = rnorm(n(), flowering_doy_mu, params$sigma), # TRUE flowering by allowing SD
    year = year)
  
  true_occurrences_all[[as.character(year)]] <- true_occurrences_year
  
}

# combine all years
true_occurrences <- bind_rows(true_occurrences_all)

true_flowering_plot_all <- ggplot() + 
  geom_sf(data = nevada, fill = "gray90", color = "black") + 
  geom_sf(data = species_range, fill = NA, color = "black", size = 1) + 
  geom_sf(data = true_occurrences_sf, mapping = aes(color = flowering_doy), size = 0.5) + 
  scale_color_viridis_c(option = "magma", direction = -1, name = "Flowering\nDOY") + 
  ggtitle("True Flowering DOY Across Species Range")

```

#### Step 3: Create a sampling layer

```{r}
# simple, make even sampling per year
n_per_year <- 250 

# sample evenly across years
observed_occs <- true_occurrences %>% 
  group_by(year) %>% 
  slice_sample(n = n_per_year, replace = FALSE) %>% 
  ungroup()


# visualize sampling 
# create a spatial plot: 
observed_occs_sf <- st_as_sf(observed_occs, 
                                coords = c("X", "Y"), 
                                crs = crs(tmean_nev))

true_flowering_plot <- ggplot() + 
  geom_sf(data = nevada, fill = "gray90", color = "black") + 
  geom_sf(data = species_range, fill = NA, color = "black", size = 1) + 
  geom_sf(data = observed_occs_sf, mapping = aes(color = flowering_doy), size = 0.5) + 
  scale_color_viridis_c(option = "magma", direction = -1, name = "Flowering\nDOY") + 
  ggtitle("True Flowering DOY Across Species Range") + 
  facet_wrap(~year)

true_flowering_plot 

ggplot() +
  geom_point(data = true_occurrences, aes(x = tmean_spring, y = flowering_doy), 
             color = "black", alpha = 0.2, size = 0.5) +
  geom_point(data = observed_occs, aes(x = tmean_spring, y = flowering_doy) ,color = "steelblue", 
             alpha = 0.6) +
  geom_line(data = response_df, aes(x = temp_range, y = flowering_doy), 
            color = "darkred", size = 1.2, alpha = 0.1) +
  labs(x = "Mean Spring Temperature (°C)",
       y = "Flowering Onset (DOY)",
       color = "Year",
       title = "Observed Occurrences (blue) vs True Population (black)") +
  theme_bw() + 
  facet_wrap(~year)

```

#### Step 4: Group clusters to build phenometric estimates

```{r}
# use dbscan to group clusters by year
cluster_list <- list()

for(i in unique(observed_occs$year)){
  # make a spatial dataset
  yr_df <- observed_occs %>% dplyr::filter(year == i)
  yr_df_sf <- yr_df %>% st_as_sf(coords = c("X", "Y"), crs = st_crs(tmean_nev))
  yr_df_sf <- yr_df_sf %>% st_transform(crs = 5070) # equal area proj Albers
  # extract coords
  coords <- st_coordinates(yr_df_sf)
  # use hierarchical dbscan, just requires num pts we need for phenometrics
  cl <- hdbscan(coords, minPts = 7)
  yr_df_sf$cluster_id <- cl$cluster # the id 
  yr_df_sf$cluster_prob <- cl$membership_prob # the clusters support val
  cluster_list[[as.character(i)]] <- yr_df_sf
}

# view clusters
plot_df <- cluster_list[["2017"]] %>% 
  mutate(cluster_id = factor(cluster_id))

ggplot(plot_df) + 
  geom_sf(aes(color = cluster_id), size = 1.5, alpha = 0.8) + 
  guides(color = guide_legend(title = "Cluster ID")) + 
  theme_bw()

## Spatial Clustering
cluster_spatial <- function(data, distance_km = 50, min_obs = 7){
  data_clustered <- data %>% 
    group_by(year) %>% 
    group_modify(~ {
      coords <- as.matrix(.x[, c("X", "Y")])
      # convert distance to meters
      eps_m <- distance_km * 1000
      # DBSCAN clustering 
      clusters <- dbscan(coords, eps = eps_m, minPts = min_obs)
      .x %>% 
        mutate(cluster_id = clusters$cluster) %>% 
        filter(cluster_id != 0) # removes pts IDed as noise
    }) %>% 
    ungroup() %>% 
    # create unique cluster ID across years
    mutate(cluster_full_id = paste0(year, "_", cluster_id)) %>% 
    # filter to clusters with minimum obs
    group_by(cluster_full_id) %>% 
    filter(n() >= min_obs) %>% 
    # calc spatial area per cluster
     mutate(
      cluster_n = n(),
      # Mean distance to cluster centroid (km) - LOWER = denser
      cluster_mean_dist_to_center = sqrt(mean((X - mean(X))^2 + (Y - mean(Y))^2)) / 1000,
      # Mean nearest neighbor distance (km) - LOWER = denser
      cluster_mean_nn_dist = {
        coords_mat <- as.matrix(data.frame(X = X, Y = Y))
        dist_mat <- as.matrix(dist(coords_mat))
        diag(dist_mat) <- NA  # Exclude self-distance
        mean(apply(dist_mat, 1, min, na.rm = TRUE)) / 1000
      },
      # Inverse for weighting (lower distance = higher weight)
      cluster_compactness = 1 / (cluster_mean_dist_to_center + 0.1)
    ) %>%
    ungroup()
  
  return(data_clustered)
}

## Environmental Clustering 
cluster_env <- function(data, temp_range_c = 2, min_obs = 7){
  data_clustered <- data %>% 
    group_by(year) %>% 
    group_modify(~{
      # run hierarchical clustering on temp
      temp_dist <- dist(.x$tmean_spring)
      hc <- hclust(temp_dist, method = "complete")
      # cut tree to get clusters with max temp_range_c difference
      clusters <- cutree(hc, h = temp_range_c)
      .x %>% 
        mutate(cluster_id = clusters)
    }) %>% 
    ungroup() %>% 
    # create unique cluster IDs across years
    mutate(cluster_full_id = paste0(year, "_", cluster_id)) %>% 
    # Filter to clusters with min obs
    group_by(cluster_full_id) %>% 
    filter(n() >= min_obs) %>% 
    # Calculate environmental variation per cluster
    mutate(
      cluster_n = n(),
      cluster_temp_sd = sd(tmean_spring),
      cluster_temp_range = max(tmean_spring) - min(tmean_spring),
      cluster_temp_cv = sd(tmean_spring) / abs(mean(tmean_spring))
    ) %>%
    ungroup()
  
  return(data_clustered)
}
obs_spatial_clusters <- cluster_spatial(observed_occs, distance_km = 25, min_obs = 7)
obs_env_clusters <- cluster_env(observed_occs, temp_range_c = 2, min_obs = 7)

# Visualize clusters in space
obs_spatial_clusters_sf <- st_as_sf(obs_spatial_clusters, 
                                     coords = c("X", "Y"), 
                                     crs = crs(tmean_nev))
obs_env_clusters_sf <- st_as_sf(obs_env_clusters, 
                                     coords = c("X", "Y"), 
                                     crs = crs(tmean_nev))

ggplot() +
  geom_sf(data = nevada, fill = "gray90") +
  geom_sf(data = obs_spatial_clusters_sf, aes(color = factor(cluster_id)), size = 1) +
  facet_wrap(~year) +
  labs(color = "Cluster", title = "Spatial Clusters by Year") +
  theme_minimal()

ggplot() +
  geom_sf(data = nevada, fill = "gray90") +
  geom_sf(data = obs_env_clusters_sf, aes(color = factor(cluster_id)), size = 1) +
  facet_wrap(~year) +
  labs(color = "Cluster", title = "Temp Env Clusters by Year") +
  theme_minimal()
```

#### Step 5: Use phenesse to Calc Onset

```{r}
cluster_spatial_onset <- obs_spatial_clusters %>%
  group_by(cluster_full_id, year, cluster_compactness) %>%
  summarise(
    n_obs = n(),
    onset_estimate = phenesse::quantile_ci(observations = flowering_doy, 
                                  percentile = 0.10, 
                                  bootstraps = 250)$estimate,
    # Also get cluster-level environmental predictors for GAM
    tmean_spring_mean = mean(tmean_spring),
    ppt_spring_mean = mean(ppt_spring),
    X_centroid = mean(X),
    Y_centroid = mean(Y),
    .groups = "drop"
  )  %>% 
  mutate(weight = cluster_compactness / mean(cluster_compactness)) # normalized

cluster_env_onset <- obs_env_clusters %>%
  group_by(cluster_full_id, year, cluster_temp_sd) %>%
  summarise(
    n_obs = n(),
    onset_estimate = phenesse::quantile_ci(observations = flowering_doy, 
                                  percentile = 0.10, 
                                  bootstraps = 250)$estimate,
    # Also get cluster-level environmental predictors for GAM
    tmean_spring_mean = mean(tmean_spring),
    ppt_spring_mean = mean(ppt_spring),
    X_centroid = mean(X),
    Y_centroid = mean(Y),
    .groups = "drop"
  )  %>% 
  mutate(weight = mean(cluster_temp_sd)/cluster_temp_sd)

```

#### Step 6. Model Onset using GAM

```{r}
cluster_spatial_onset$cluster_full_id <- as.factor(cluster_spatial_onset$cluster_full_id)
cluster_env_onset$cluster_full_id <- as.factor(cluster_env_onset$cluster_full_id)
# run GAM, overkill here.
gam_spatial <- gam(onset_estimate ~ s(tmean_spring_mean, k = 5) + 
                     s(cluster_full_id, bs = "re"), 
                   data = cluster_spatial_onset, 
                   weights = weight)

gam_env <- gam(onset_estimate ~ s(tmean_spring_mean, k = 5) + 
                     s(cluster_full_id, bs = "re"), 
                   data = cluster_env_onset, 
                   weights = weight)

# predict 
true_occurrences$predicted_onset_spatial <- predict(gam_spatial, newdata = data.frame(
                        tmean_spring_mean = true_occurrences$tmean_spring), 
                        exclude = "s(cluster_full_id)",  # Exclude random effect
                        newdata.guaranteed = TRUE)

true_occurrences$predicted_onset_env <- predict(gam_env, newdata = data.frame(
                        tmean_spring_mean = true_occurrences$tmean_spring), 
                        exclude = "s(cluster_full_id)",  # Exclude random effect
                        newdata.guaranteed = TRUE)


```
