---
title: "simu-data-v3"
author: "JT Miller"
format: html
editor: visual
---

## Simulate Pheno Extraction and Interpolation Across a species Range

### Methods

Load Packages

```{r}
library(rnaturalearth)
library(sf)
library(ggplot2)
library(terra)
library(dplyr)
library(patchwork)
library(dbscan)
library(mgcv)
```

#### Step 1: Create Nevada Phenology Simulation with Daymet climate Data

```{r}
# get nevada boundary
nevada <- ne_states(country = "United States of America", 
                    returnclass = "sf") %>% filter(name == "Nevada")

```

```{r}
#| eval: false
# Path to Daymet data
# daymet_path <- "/orange/guralnick/daymet-daily/"
# 
# # Function to extract Nevada data from Daymet NetCDF
# extract_nevada_daymet <- function(variable, years) {
#   
#   climate_list <- list()
#   
#   for(year in years) {
#     # Load NetCDF file
#     nc_file <- file.path(daymet_path, 
#                          paste0("daymet_v4_daily_na_", variable, "_", year, ".nc"))
#     climate_rast <- rast(nc_file)
#     
#     # Reproject Nevada to match Daymet CRS
#     nevada_reproj <- st_transform(nevada, crs(climate_rast))
#     nevada_vect <- vect(nevada_reproj)
#     
#     # Crop and mask to Nevada
#     climate_nv <- crop(climate_rast, nevada_vect, mask = TRUE)
#     
#     climate_list[[as.character(year)]] <- climate_nv
#   }
#   
#   # Combine all years
#   climate_all <- do.call(c, climate_list)
#   
#   return(climate_all)
# }
# 
# # Extract data for 2017-2020
# years <- 2017:2020
# tmin_nv <- extract_nevada_daymet("tmin", years)
# tmax_nv <- extract_nevada_daymet("tmax", years)
# prcp_nv <- extract_nevada_daymet("prcp", years)
# 
# # combine elements into a singular raster
# tmin_nv_comined <- rast(tmin_nv)
# tmax_nv_comined <- rast(tmax_nv)
# prcp_nv_comined <- rast(prcp_nv)
# 
# # additionally calc tmean from tmin and tmax 
# tmean_nv <- (tmin_nv_comined + tmax_nv_comined)/2
# 
# # Save for later use
# writeRaster(tmin_nv_comined, "/blue/guralnick/millerjared/PhenoInterpolate/data/processed/clim-layers/nevada_tmin_2017_2020_daymet.tif", overwrite = TRUE)
# writeRaster(tmax_nv_comined, "/blue/guralnick/millerjared/PhenoInterpolate/data/processed/clim-layers/nevada_tmax_2017_2020_daymet.tif", overwrite = TRUE)
# writeRaster(prcp_nv_comined, "/blue/guralnick/millerjared/PhenoInterpolate/data/processed/clim-layers/nevada_prcp_2017_2020_daymet.tif", overwrite = TRUE)
# writeRaster(tmean_nv, "/blue/guralnick/millerjared/PhenoInterpolate/data/processed/clim-layers/nevada_tmean_2017_2020_daymet.tif", overwrite = TRUE)

```

Read in Raster Stack

```{r}
tmin_nev <- terra::rast("/blue/guralnick/millerjared/PhenoInterpolate/data/processed/clim-layers/nevada_tmin_2017_2020_daymet.tif")
tmax_nev <- terra::rast("/blue/guralnick/millerjared/PhenoInterpolate/data/processed/clim-layers/nevada_tmin_2017_2020_daymet.tif")
tmean_nev <- terra::rast("/blue/guralnick/millerjared/PhenoInterpolate/data/processed/clim-layers/nevada_tmean_2017_2020_daymet.tif")
ppt_nev <- terra::rast("/blue/guralnick/millerjared/PhenoInterpolate/data/processed/clim-layers/nevada_prcp_2017_2020_daymet.tif")
```

Create a fake species range

```{r}
nevada_bbox <- st_bbox(nevada)

species_range <- st_polygon(list(cbind(
                        c(-120, -119, -117, -115, -116, -118, -120), # lons
                        c(36, 39, 41, 39.5, 37, 35.5, 36)))) %>% #lats
  st_sfc(crs = st_crs(nevada)) %>% 
  st_sf() %>% 
  st_intersection(nevada) # clip to nev boundary

ggplot() + 
  geom_sf(data = nevada, mapping = aes(), fill = "lightgray") + 
  geom_sf(data = species_range, mapping = aes(), fill = "darkgreen", alpha = 0.5) + 
  theme_bw() + 
  ggtitle("Species S Distribution in Nevada")
```

#### Step 2: Set-up flowering Response

```{r}
#| output: true
## Create a true data generating process
params <- list(
  intercept = 170,      # flowering DOY when temp = 0°C
  temp_slope = -3.5,    # days earlier per 1°C increase
  sigma_spatial = 5, # variation between sites
  sigma_individual = 2, # variation within populations
  ppt_min = 25, # min ppt mm needed for flower on temp schedule
  ppt_max = 85, # ppt above which penalty on flowering begins
  ppt_max_penalty = 60 # maximum DOY penalty for excess ppt
)

# Population flowering distributions
sim_pop_flowering <- function(temp, ppt, n_individuals = 100, par = params){
  mu_site <- par$intercept + par$temp_slope * temp
  
  ppt_mid <- (par$ppt_min + par$ppt_max) / 2  # peak of parabola (minimum penalty)
  
  ppt_penalty <- if(ppt < par$ppt_min){
    par$ppt_max_penalty  # below minimum: flat max penalty
  } else if(ppt > par$ppt_max){
    par$ppt_max_penalty  # above maximum: flat max penalty
  } else {
    # parabola between ppt_min and ppt_max, 0 at midpoint, ppt_max_penalty at edges
    par$ppt_max_penalty * ((ppt - ppt_mid) / (ppt_mid - par$ppt_min))^2
  }
  
  mu_site <- mu_site + ppt_penalty
  
  flowering_times <- rnorm(n_individuals,
                           mean = mu_site,
                           sd = par$sigma_individual)
  return(flowering_times)
}
# visualize response curve with phenology dists
ppt_scenarios <- c(20, 30, 45, 100)  # drought, adequate, optimal, excess
ppt_labels <- c("Drought (<25mm)", "Adequate (30mm)", "Optimal (50mm)", "Excess (>75mm)")
temp_range <- seq(0, 25, by = 0.5)

response_distributions <- expand.grid(temp_range = temp_range, ppt = ppt_scenarios) %>%
  mutate(ppt_label = factor(ppt, levels = ppt_scenarios, labels = ppt_labels)) %>%
  mutate(
    distribution = purrr::pmap(list(temp_range, ppt), function(t, p)
      sim_pop_flowering(t, p, n_individuals = 500, par = params)),
    onset_10  = purrr::map_dbl(distribution, ~quantile(.x, 0.10)),
    median_50 = purrr::map_dbl(distribution, ~quantile(.x, 0.50)),
    offset_90 = purrr::map_dbl(distribution, ~quantile(.x, 0.90))
  )


response_long <- response_distributions %>%
  tidyr::pivot_longer(cols = c(onset_10, median_50, offset_90),
                      names_to = "phenometric",
                      values_to = "doy") %>%
  mutate(phenometric = factor(phenometric,
                              levels = c("onset_10", "median_50", "offset_90"),
                              labels = c("Onset (10%)", "Median (50%)", "Offset (90%)")))

ggplot(response_long, aes(x = temp_range, y = doy, color = phenometric, linetype = phenometric)) +
  geom_ribbon(data = response_distributions,
              aes(x = temp_range, ymin = onset_10, ymax = offset_90),
              fill = "yellow", alpha = 0.2, inherit.aes = FALSE) +
  geom_line(size = 1) +
  scale_color_manual(values = c("darkred", "darkorange", "steelblue")) +
  scale_linetype_manual(values = c("dashed", "solid", "dashed")) +
  facet_wrap(~ppt_label) +
  labs(color = "Phenometric", linetype = "Phenometric",
       x = "Mean Spring Temperature (°C)", y = "Flowering DOY",
       title = "Flowering Response to Temperature Across Precipitation Scenarios") +
  theme_bw()
```

Match up species range with climate, then simu spring flowering

```{r}
#| output: true
# reproj species range to match clim data
species_range_clim <- st_transform(species_range, crs = crs(tmean_nev))

# create a point for each 1km cell center within the species range (res of daymet clim layer)
tmean_species <- crop(tmean_nev, vect(species_range_clim), mask = TRUE)
ppt_species <- crop(ppt_nev, vect(species_range_clim), mask = TRUE)
# extract cell centers as pts 
cells <- cells(tmean_species[[1]]) # get cell numbers
cell_coords <- xyFromCell(tmean_species[[1]], cells)
cell_pts <- st_as_sf(data.frame(cell_coords), 
                    coords = c("x", "y"), 
                    crs = crs(tmean_nev))

## Do this for each year 
years <- 2017:2020
days_per_year <- c(365,365,365,366)
spring_doy <- 30:150

# calc cumulative yera positions 
year_starts <- c(1, cumsum(days_per_year[-length(days_per_year)]) + 1)
names(year_starts) <- years

# genearte true occurrences for all years
cells_true_flowering_all <- list()
for(i in seq_along(years)){
  # get year and when year starts in raster
  year <- years[i]
  year_start <- year_starts[i]
  # get spring layers for this year
  spring_layers_year <- (year_start - 1) + spring_doy
  # calc spring climate for this year
  tmean_spring <- mean(tmean_species[[spring_layers_year]])
  ppt_spring <- sum(ppt_species[[spring_layers_year]])
  # extract clim vals 
  climate_data <- terra::extract(c(tmean_spring, ppt_spring), vect(cell_pts))
  # generate cell phenometrics per year
  cells_true_flowering__dists_year <- cell_pts %>% 
  st_drop_geometry() %>% 
  bind_cols(st_coordinates(cell_pts)) %>% 
  mutate(
    tmean_spring = climate_data$mean,
    ppt_spring = climate_data$sum, 
    year = year
  ) %>% 
  rowwise() %>% 
  mutate(
    cell_flowering_distribution = list(sim_pop_flowering(tmean_spring, ppt = ppt_spring, n_individuals = 100, par = params)
  ), 
  cell_flowering_onset = quantile(cell_flowering_distribution, 0.10), 
  cell_flowering_median = quantile(cell_flowering_distribution, 0.50), 
  cell_flowering_offset = quantile(cell_flowering_distribution, 0.90)) %>% 
  ungroup() 
  cells_true_flowering_all[[as.character(year)]] <- cells_true_flowering__dists_year
 
}

# combine all years
cells_true_flowering <- bind_rows(cells_true_flowering_all)
cells_true_flowering_sf  <- st_as_sf(cells_true_flowering, 
                                coords = c("X", "Y"), 
                                crs = crs(tmean_nev))


true_flowering_plot_all <- ggplot() + 
  geom_sf(data = nevada, fill = "gray90", color = "black") + 
  geom_sf(data = species_range, fill = NA, color = "black", size = 1) + 
  geom_sf(data = cells_true_flowering_sf , mapping = aes(color = cell_flowering_onset), size = 0.5) + 
  scale_color_viridis_c(option = "magma", direction = -1, name = "Flowering\nDOY") + 
  facet_wrap(~year) +
  ggtitle("True Flowering Onset (10%) Across Species Range") +
  theme_bw()
  
print(true_flowering_plot_all)

## create a vis to understand more about the species range. 
# Storage for results
spring_tmean_all <- c()
spring_ppt_all   <- c()

for (i in seq_along(years)) {
  yr <- years[i]
  
  # Identify spring layers for this year
  spring_layers_year <- (year_starts[i] + spring_doy - 1)
  
  # Mean spring temp and total spring ppt per cell
  tmean_spring_r <- mean(tmean_species[[spring_layers_year]])
  ppt_spring_r   <- sum(ppt_species[[spring_layers_year]])
  
  # Extract values at cell centers
  tmean_vals <- terra::extract(tmean_spring_r, vect(cell_pts))[, 2]
  ppt_vals   <- terra::extract(ppt_spring_r,   vect(cell_pts))[, 2]
  
  spring_tmean_all <- c(spring_tmean_all, tmean_vals)
  spring_ppt_all   <- c(spring_ppt_all,   ppt_vals)
}

# Combine into a data frame, dropping NAs
df <- data.frame(
  tmean_spring = spring_tmean_all,
  ppt_spring   = spring_ppt_all
) |> na.omit()

# Helper to compute summary stats
summary_stats <- function(x) {
  data.frame(
    mean = mean(x),
    p05  = quantile(x, 0.05),
    p95  = quantile(x, 0.95)
  )
}

stats_tmean <- summary_stats(df$tmean_spring)
stats_ppt   <- summary_stats(df$ppt_spring)

# --- Plot: Spring Tmean ---
p1 <- ggplot(df, aes(x = tmean_spring)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 50, fill = "steelblue", alpha = 0.7, color = "white") +
  geom_vline(xintercept = stats_tmean$mean, color = "black",  linewidth = 1,   linetype = "solid",  label = "Mean") +
  geom_vline(xintercept = stats_tmean$p05,  color = "firebrick", linewidth = 0.8, linetype = "dashed") +
  geom_vline(xintercept = stats_tmean$p95,  color = "firebrick", linewidth = 0.8, linetype = "dashed") +
  annotate("text", x = stats_tmean$mean, y = Inf, label = paste0("Mean: ", round(stats_tmean$mean, 1)),
           vjust = 2, hjust = -0.1, size = 3.5) +
  annotate("text", x = stats_tmean$p05, y = Inf, label = "5th",
           vjust = 2, hjust = -0.1, color = "firebrick", size = 3.5) +
  annotate("text", x = stats_tmean$p95, y = Inf, label = "95th",
           vjust = 2, hjust = -0.1, color = "firebrick", size = 3.5) +
  labs(title = "Distribution of Mean Spring Temperature",
       x = "Mean Spring Tmean (°C)", y = "Density") +
  theme_classic()

# --- Plot: Spring PPT ---
p2 <- ggplot(df, aes(x = ppt_spring)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 50, fill = "darkgreen", alpha = 0.7, color = "white") +
  geom_vline(xintercept = stats_ppt$mean, color = "black",     linewidth = 1,   linetype = "solid") +
  geom_vline(xintercept = stats_ppt$p05,  color = "firebrick", linewidth = 0.8, linetype = "dashed") +
  geom_vline(xintercept = stats_ppt$p95,  color = "firebrick", linewidth = 0.8, linetype = "dashed") +
  annotate("text", x = stats_ppt$mean, y = Inf, label = paste0("Mean: ", round(stats_ppt$mean, 1)),
           vjust = 2, hjust = -0.1, size = 3.5) +
  annotate("text", x = stats_ppt$p05, y = Inf, label = "5th",
           vjust = 2, hjust = -0.1, color = "firebrick", size = 3.5) +
  annotate("text", x = stats_ppt$p95, y = Inf, label = "95th",
           vjust = 2, hjust = -0.1, color = "firebrick", size = 3.5) +
  labs(title = "Distribution of Total Spring Precipitation",
       x = "Total Spring PPT (mm)", y = "Density") +
  theme_classic()

print(p1)
print(p2)

```

#### Step 3: Create a sampling layer

```{r}
#| output: true
# start simple, lets evenly sample cells across space and across flowering distributions
n_cells_per_year <- 50
obs_per_cell_range <- c(5,20)

# sample from true pops 
observed_occs <- cells_true_flowering %>% 
  group_by(year) %>% 
  # randomly sample cells 
  slice_sample(n = n_cells_per_year, replace = FALSE) %>% 
  ungroup() %>% 
  rowwise() %>% 
  mutate( # randomly determine how many obs within cell
    n_obs_at_cell = sample(obs_per_cell_range[1]:obs_per_cell_range[2], 1), 
    # sample flowering from distribution
    observed_flowering_times = list(sample(cell_flowering_distribution, n_obs_at_cell))
  ) %>% 
  ungroup() %>% 
  tidyr::unnest(observed_flowering_times) %>%  # each row will be an obs
  rename(flowering_doy_obs = observed_flowering_times)

# visualize sampling 
# create a spatial plot: 
observed_occs_sf <- st_as_sf(observed_occs, 
                                coords = c("X", "Y"), 
                                crs = crs(tmean_nev))

observed_occs_sf <- observed_occs_sf %>% 
  distinct(geometry, year, n_obs_at_cell)

obs_flowering_plot <- ggplot() + 
  geom_sf(data = nevada, fill = "gray90", color = "black") + 
  geom_sf(data = species_range, fill = NA, color = "black", size = 1) + 
  geom_sf(data = observed_occs_sf, mapping = aes(color = n_obs_at_cell), size = 0.5) + 
  scale_color_viridis_c(option = "magma", direction = +1, name = "Number of Obs") + 
  ggtitle("Flowering Observation Simu") + 
  facet_wrap(~year)

obs_flowering_plot 

```

#### Step 4: Group clusters to build phenometric estimates

```{r}
#| output: true
## Spatial Clustering
cluster_spatial <- function(data, distance_km = 50, min_obs = 7, min_cells = 3){
  data_clustered <- data %>% 
    group_by(year) %>% 
    group_modify(~{
      # get unique cell locs for clustering
      unique_cells <- .x %>% 
        distinct(X,Y, .keep_all = TRUE) %>% 
        dplyr::select(X,Y, tmean_spring, ppt_spring)
      coords <- as.matrix(unique_cells[, c("X", "Y")])
      # convert dist to meters
      eps_m <- distance_km * 1000
      # DBSCAN clustering on cells, instead of individual obs
      clusters <- dbscan(coords, eps = eps_m, minPts = min_cells)
      # add cluster IDs to unique cells
      unique_cells$cluster_id <- clusters$cluster
      # join cluster IDs back to all obs
      .x %>% 
        left_join(unique_cells %>% dplyr::select(X,Y,cluster_id), by = c("X","Y")) %>% 
        filter(cluster_id != 0) # removes noise cluster
    }) %>% 
    ungroup() %>% 
    # create unique cluster id across years
    mutate(cluster_full_id = paste0(year, "_", cluster_id)) %>% 
    # filter to clusters with min total obs
    group_by(cluster_full_id) %>% 
    filter(n() >= min_obs) %>% 
    # calc spatial metrics per cluster
    mutate(
      cluster_n_obs = n(), # total obs in cluster
      cluster_n_cells = n_distinct(X,Y), # number of cells in cluster
      # Calc cell level centroids for spatial metrics 
      cluster_mean_dist_to_center = {
        cell_coords <- distinct(pick(X,Y))
        centroid_x <- mean(cell_coords$X)
        centroid_y <- mean(cell_coords$Y)
        sqrt(mean((cell_coords$X - centroid_x)^2 + (cell_coords$Y - centroid_y)^2)) / 1000
      }, 
      # mean nearest neighbor dist between cells (km)
      cluster_mean_nn_dist = {
        cell_coords <- distinct(pick(X, Y))
        if(nrow(cell_coords) > 1) {
          coords_mat <- as.matrix(cell_coords)
          dist_mat <- as.matrix(dist(coords_mat))
          diag(dist_mat) <- NA
          mean(apply(dist_mat, 1, min, na.rm = TRUE)) / 1000
        } else {
          NA_real_
        }
      },
      
      # Inverse for weighting (lower distance = higher weight)
      cluster_compactness = 1 / (cluster_mean_dist_to_center + 0.1)
    ) %>%
    ungroup()
  
  return(data_clustered)
}


## Environmental Clustering 
 cluster_env <- function(data, env_range_h = 2, min_obs = 7, min_cells = 3){
  data_clustered <- data %>%
    group_by(year) %>%
    group_modify(~{
      # get unique cell locations with their env vars
      unique_cells <- .x %>%
        distinct(X, Y, .keep_all = TRUE) %>%
        dplyr::select(X, Y, tmean_spring, ppt_spring)
      
      # scale both variables so neither dominates by units
      env_scaled <- scale(unique_cells[, c("tmean_spring", "ppt_spring")])
      
      # run hierarchical clustering on scaled env space
      env_dist <- dist(env_scaled)
      hc <- hclust(env_dist, method = "complete")
      clusters <- cutree(hc, h = env_range_h)
      
      unique_cells$cluster_id <- clusters
      
      .x %>%
        left_join(unique_cells %>% dplyr::select(X, Y, cluster_id), by = c("X", "Y"))
    }) %>%
    ungroup() %>%
    mutate(cluster_full_id = paste0(year, "_", cluster_id)) %>%
    group_by(cluster_full_id) %>%
    filter(n_distinct(X, Y) >= min_cells,
           n() >= min_obs) %>%
    mutate(
      cluster_n_obs   = n(),
      cluster_n_cells = n_distinct(X, Y),
      # Temperature metrics
      cluster_temp_mean  = {
        cell_temps <- distinct(pick(X, Y, tmean_spring))$tmean_spring
        mean(cell_temps)
      },
      cluster_temp_sd  = {
        cell_temps <- distinct(pick(X, Y, tmean_spring))$tmean_spring
        sd(cell_temps)
      },
      cluster_temp_range = {
        cell_temps <- distinct(pick(X, Y, tmean_spring))$tmean_spring
        max(cell_temps) - min(cell_temps)
      },
      # Precipitation metrics
      cluster_ppt_mean = {
        cell_ppts <- distinct(pick(X, Y, ppt_spring))$ppt_spring
        mean(cell_ppts)
      },
      cluster_ppt_sd = {
        cell_ppts <- distinct(pick(X, Y, ppt_spring))$ppt_spring
        sd(cell_ppts)
      },
      cluster_ppt_range = {
        cell_ppts <- distinct(pick(X, Y, ppt_spring))$ppt_spring
        max(cell_ppts) - min(cell_ppts)
      },
      # Weight: prefer homogeneous clusters in both dimensions
      # combines temp and ppt sd so clusters tight in both get higher weight
      cluster_env_weight = 1 / (cluster_temp_sd + cluster_ppt_sd + 0.1)
    ) %>%
    ungroup()
  return(data_clustered)
 }
 
obs_spatial_clusters <- cluster_spatial(observed_occs, 
                                        distance_km = 25, 
                                        min_obs = 7,     
                                        min_cells = 2)    
obs_env_clusters <- cluster_env(observed_occs, 
                                env_range_h = 2, 
                                min_obs = 7, 
                                min_cells = 2)

# Visualize clusters in space
obs_spatial_clusters_sf <- st_as_sf(obs_spatial_clusters, 
                                     coords = c("X", "Y"), 
                                     crs = crs(tmean_nev))
obs_env_clusters_sf <- st_as_sf(obs_env_clusters, 
                                     coords = c("X", "Y"), 
                                     crs = crs(tmean_nev))

ggplot() +
  geom_sf(data = nevada, fill = "gray90") +
  geom_sf(data = obs_spatial_clusters_sf, aes(color = factor(cluster_id)), size = 1) +
  facet_wrap(~year) +
  labs(color = "Cluster", title = "Spatial Clusters by Year") +
  theme_minimal()

ggplot() +
  geom_sf(data = nevada, fill = "gray90") +
  geom_sf(data = obs_env_clusters_sf, aes(color = factor(cluster_id)), size = 1) +
  facet_wrap(~year) +
  labs(color = "Cluster", title = "Temp Env Clusters by Year") +
  theme_minimal()
```

#### Step 5: Use phenesse to Calc Onset

```{r}
#| output: true
cluster_spatial_onset <- obs_spatial_clusters %>%
  group_by(cluster_full_id, year, cluster_compactness) %>%
  summarise(
    n_obs = n(),
    onset_estimate = phenesse::quantile_ci(observations = flowering_doy_obs, 
                                  percentile = 0.10, 
                                  bootstraps = 250)$estimate,
    # Also get cluster-level environmental predictors for GAM
    tmean_spring_mean = mean(tmean_spring),
    ppt_spring_mean = mean(ppt_spring),
    X_centroid = mean(X),
    Y_centroid = mean(Y),
    .groups = "drop"
  )  %>% 
  mutate(weight = cluster_compactness / mean(cluster_compactness)) # normalized

cluster_env_onset <- obs_env_clusters %>%
  group_by(cluster_full_id, year, cluster_temp_sd) %>%
  summarise(
    n_obs = n(),
    onset_estimate = phenesse::quantile_ci(observations = flowering_doy_obs, 
                                  percentile = 0.10, 
                                  bootstraps = 250)$estimate,
    # Also get cluster-level environmental predictors for GAM
    tmean_spring_mean = mean(tmean_spring),
    ppt_spring_mean = mean(ppt_spring),
    X_centroid = mean(X),
    Y_centroid = mean(Y),
    .groups = "drop"
  )  %>% 
  mutate(weight = mean(cluster_temp_sd)/cluster_temp_sd)

```

#### Step 6. Model Onset using GAM

```{r}
#| output: true
cluster_spatial_onset$cluster_full_id <- as.factor(cluster_spatial_onset$cluster_full_id)
cluster_env_onset$cluster_full_id <- as.factor(cluster_env_onset$cluster_full_id)
# run GAM, overkill here.
gam_spatial <- gam(onset_estimate ~ s(tmean_spring_mean, k = 5) +
                     s(ppt_spring_mean, k = 7) +
                     s(cluster_full_id, bs = "re"), 
                   data = cluster_spatial_onset, 
                   weights = weight)

gam_env <- gam(onset_estimate ~ s(tmean_spring_mean, k = 7) + 
                     s(ppt_spring_mean, k = 5) +
                     s(cluster_full_id, bs = "re"), 
                   data = cluster_env_onset, 
                   weights = weight)

# predict 
cells_true_flowering$predicted_onset_spatial <- predict(gam_spatial, newdata = data.frame(
                        tmean_spring_mean = cells_true_flowering$tmean_spring,
                        ppt_spring_mean = cells_true_flowering$ppt_spring), 
                        exclude = "s(cluster_full_id)",  # Exclude random effect
                        newdata.guaranteed = TRUE)

cells_true_flowering$predicted_onset_env <- predict(gam_env, newdata = data.frame(
                        tmean_spring_mean = cells_true_flowering$tmean_spring, 
                        ppt_spring_mean = cells_true_flowering$ppt_spring), 
                        exclude = "s(cluster_full_id)",  # Exclude random effect
                        newdata.guaranteed = TRUE)


```

#### Step 7. Calculate MESS metrics

```{r}
#| output: true
# predictor vars to check 
predictor_vars <- c("tmean_spring", "ppt_spring") 

# set-up training data (empirical data) for the models
training_cells_spatial <- obs_spatial_clusters %>% 
  distinct(X, Y, .keep_all = TRUE)
training_cells_env <- obs_env_clusters %>% 
  distinct(X, Y, .keep_all = TRUE)

# MESS per var

for(v in 1:length(predictor_vars)){
  var_name <- predictor_vars[v]
  
  # Spatial
  low  <- min(training_cells_spatial[[var_name]], na.rm = TRUE)
  high <- max(training_cells_spatial[[var_name]], na.rm = TRUE)
  cells_true_flowering[[paste0("mess_spatial_", var_name)]] <-
    ifelse(cells_true_flowering[[var_name]] < low | cells_true_flowering[[var_name]] > high, 0, 1)
  cells_true_flowering[[paste0("mess_dist_spatial_", var_name)]] <-
    ifelse(cells_true_flowering[[paste0("mess_spatial_", var_name)]] == 1, 0,
           pmin(abs(cells_true_flowering[[var_name]] - low),
                abs(cells_true_flowering[[var_name]] - high)))
  
  # Environmental
  low  <- min(training_cells_env[[var_name]], na.rm = TRUE)
  high <- max(training_cells_env[[var_name]], na.rm = TRUE)
  cells_true_flowering[[paste0("mess_env_", var_name)]] <-
    ifelse(cells_true_flowering[[var_name]] < low | cells_true_flowering[[var_name]] > high, 0, 1)
  cells_true_flowering[[paste0("mess_dist_env_", var_name)]] <-
    ifelse(cells_true_flowering[[paste0("mess_env_", var_name)]] == 1, 0,
           pmin(abs(cells_true_flowering[[var_name]] - low),
                abs(cells_true_flowering[[var_name]] - high)))
}

## Combine MESS interpolations only if in range of both vars
cells_true_flowering <- cells_true_flowering %>%
  mutate(
    # 1 only if interpolating in both temp and ppt
    mess_spatial_combined = as.integer(mess_spatial_tmean_spring == 1 & mess_spatial_ppt_spring == 1),
    mess_env_combined     = as.integer(mess_env_tmean_spring     == 1 & mess_env_ppt_spring     == 1),
    # total extrapolation distance as euclidean distance across both vars
    mess_dist_spatial_combined = sqrt(mess_dist_spatial_tmean_spring^2 + mess_dist_spatial_ppt_spring^2),
    mess_dist_env_combined     = sqrt(mess_dist_env_tmean_spring^2     + mess_dist_env_ppt_spring^2)
  )

## Plot fxn
plot_mess <- function(mess_sf, dist_col, title, file){
  max_val <- max(mess_sf[[dist_col]], na.rm = TRUE)
  p <- ggplot() +
    geom_sf(data = nevada, fill = "gray90", color = "black") +
    geom_sf(data = species_range, fill = NA, color = "black", size = 1) +
    geom_sf(data = mess_sf, mapping = aes(color = .data[[dist_col]]), size = 0.5) +
    scale_color_gradientn(
      colors = c("darkgreen", "yellow", "darkred"),
      values = scales::rescale(c(0, 0.001, max_val)),
      name = "Extrapolation\nDistance"
    ) +
    facet_wrap(~year) +
    ggtitle(title) +
    theme_bw()
  print(p)
  ggsave(plot = p, file = file, width = 15, height = 15)
}

mess_sf <- st_as_sf(cells_true_flowering, coords = c("X","Y"), crs = crs(tmean_nev))

# Spatial model plots
plot_mess(mess_sf, "mess_dist_spatial_tmean_spring",
          "MESS: tmean_spring (Spatial Clustering)",
          "/blue/guralnick/millerjared/PhenoInterpolate/outputs/figures/MESS-spatial-tmean.png")

plot_mess(mess_sf, "mess_dist_spatial_ppt_spring",
          "MESS: ppt_spring (Spatial Clustering)",
          "/blue/guralnick/millerjared/PhenoInterpolate/outputs/figures/MESS-spatial-ppt.png")

plot_mess(mess_sf, "mess_dist_spatial_combined",
          "MESS: Combined (Spatial Clustering)",
          "/blue/guralnick/millerjared/PhenoInterpolate/outputs/figures/MESS-spatial-combined.png")

# Environmental model plots
plot_mess(mess_sf, "mess_dist_env_tmean_spring",
          "MESS: tmean_spring (Environmental Clustering)",
          "/blue/guralnick/millerjared/PhenoInterpolate/outputs/figures/MESS-env-tmean.png")

plot_mess(mess_sf, "mess_dist_env_ppt_spring",
          "MESS: ppt_spring (Environmental Clustering)",
          "/blue/guralnick/millerjared/PhenoInterpolate/outputs/figures/MESS-env-ppt.png")

plot_mess(mess_sf, "mess_dist_env_combined",
          "MESS: Combined (Environmental Clustering)",
          "/blue/guralnick/millerjared/PhenoInterpolate/outputs/figures/MESS-env-combined.png")
```

#### Step 8. Calc Preformance Metrics

```{r}
#| output: true
#### Step 8: Error Analysis by Interpolation/Extrapolation Status

# --- Calculate absolute error for each model ---
cells_true_flowering <- cells_true_flowering %>%
  mutate(
    error_spatial = predicted_onset_spatial - cell_flowering_onset,
    error_env     = predicted_onset_env     - cell_flowering_onset,
    abs_error_spatial = abs(error_spatial),
    abs_error_env     = abs(error_env),
    # Use combined status since both vars drive the model
    interp_status_spatial = ifelse(mess_spatial_combined == 1, "Interpolation", "Extrapolation"),
    interp_status_env     = ifelse(mess_env_combined     == 1, "Interpolation", "Extrapolation")
  )

# --- Plot 1: Absolute error by interpolation status (boxplot) ---
error_long <- cells_true_flowering %>%
  dplyr::select(year, abs_error_spatial, abs_error_env,
                interp_status_spatial, interp_status_env) %>%
  tidyr::pivot_longer(
    cols = c(abs_error_spatial, abs_error_env),
    names_to = "model",
    values_to = "abs_error"
  ) %>%
  mutate(
    interp_status = ifelse(model == "abs_error_spatial", interp_status_spatial, interp_status_env),
    model = recode(model,
                   "abs_error_spatial" = "Spatial Clustering",
                   "abs_error_env"     = "Environmental Clustering")
  )

error_box_plot <- ggplot(error_long, aes(x = interp_status, y = abs_error, fill = interp_status)) +
  geom_boxplot(outlier.alpha = 0.3) +
  scale_fill_manual(values = c("Interpolation" = "darkgreen", "Extrapolation" = "darkred")) +
  facet_grid(model ~ year) +
  labs(x = NULL, y = "Absolute Error (DOY)",
       title = "Prediction Error: Interpolation vs. Extrapolation (Combined MESS)",
       fill = "Status") +
  theme_bw()
print(error_box_plot)
ggsave(plot = error_box_plot,
       file = "/blue/guralnick/millerjared/PhenoInterpolate/outputs/figures/error-by-interp-status.png",
       width = 15, height = 8)

# --- Plot 2: Response curves for temp, colored by combined interp status ---
true_curve <- data.frame(tmean_spring = seq(
  min(cells_true_flowering$tmean_spring, na.rm = TRUE),
  max(cells_true_flowering$tmean_spring, na.rm = TRUE),
  length.out = 200
)) %>%
  mutate(true_onset = qnorm(0.10,
                            mean = params$intercept + params$temp_slope * tmean_spring,
                            sd = params$sigma_individual))

response_curve_spatial <- ggplot(cells_true_flowering,
                                  aes(x = tmean_spring, y = predicted_onset_spatial,
                                      color = interp_status_spatial)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_line(data = true_curve, aes(x = tmean_spring, y = true_onset),
            color = "black", linewidth = 1.2, inherit.aes = FALSE) +
  scale_color_manual(values = c("Interpolation" = "darkgreen", "Extrapolation" = "darkred")) +
  facet_wrap(~year) +
  labs(x = "Mean Spring Temperature (°C)", y = "Flowering Onset (DOY)",
       title = "Spatial Model: Predicted vs True Response (temp)",
       color = "Status") +
  theme_bw()

response_curve_env <- ggplot(cells_true_flowering,
                              aes(x = tmean_spring, y = predicted_onset_env,
                                  color = interp_status_env)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_line(data = true_curve, aes(x = tmean_spring, y = true_onset),
            color = "black", linewidth = 1.2, inherit.aes = FALSE) +
  scale_color_manual(values = c("Interpolation" = "darkgreen", "Extrapolation" = "darkred")) +
  facet_wrap(~year) +
  labs(x = "Mean Spring Temperature (°C)", y = "Flowering Onset (DOY)",
       title = "Environmental Model: Predicted vs True Response (temp)",
       color = "Status") +
  theme_bw()

# --- Plot 3: ppt response curves ---
# true ppt effect at median temp
median_temp <- median(cells_true_flowering$tmean_spring, na.rm = TRUE)
true_curve_ppt <- data.frame(ppt_spring = seq(
  min(cells_true_flowering$ppt_spring, na.rm = TRUE),
  max(cells_true_flowering$ppt_spring, na.rm = TRUE),
  length.out = 200
)) %>%
  rowwise() %>%
  mutate(
    ppt_penalty = if(ppt_spring < params$ppt_min){
      params$ppt_max_penalty
    } else if(ppt_spring > params$ppt_max){
      min(((ppt_spring - params$ppt_max) / params$ppt_max) * params$ppt_max_penalty,
          params$ppt_max_penalty)
    } else {
      0
    },
    true_onset = qnorm(0.10,
                       mean = params$intercept + params$temp_slope * median_temp + ppt_penalty,
                       sd = params$sigma_individual)
  ) %>%
  ungroup()

ppt_curve_spatial <- ggplot(cells_true_flowering,
                             aes(x = ppt_spring, y = predicted_onset_spatial,
                                 color = interp_status_spatial)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_line(data = true_curve_ppt, aes(x = ppt_spring, y = true_onset),
            color = "black", linewidth = 1.2, inherit.aes = FALSE) +
  scale_color_manual(values = c("Interpolation" = "darkgreen", "Extrapolation" = "darkred")) +
  facet_wrap(~year) +
  labs(x = "Cumulative Spring PPT (mm)", y = "Flowering Onset (DOY)",
       title = "Spatial Model: Predicted vs True Response (ppt)",
       color = "Status") +
  theme_bw()

ppt_curve_env <- ggplot(cells_true_flowering,
                         aes(x = ppt_spring, y = predicted_onset_env,
                             color = interp_status_env)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_line(data = true_curve_ppt, aes(x = ppt_spring, y = true_onset),
            color = "black", linewidth = 1.2, inherit.aes = FALSE) +
  scale_color_manual(values = c("Interpolation" = "darkgreen", "Extrapolation" = "darkred")) +
  facet_wrap(~year) +
  labs(x = "Cumulative Spring PPT (mm)", y = "Flowering Onset (DOY)",
       title = "Environmental Model: Predicted vs True Response (ppt)",
       color = "Status") +
  theme_bw()

combined_response <- (response_curve_spatial / response_curve_env) |
                     (ppt_curve_spatial / ppt_curve_env)
ggsave(plot = combined_response,
       file = "/blue/guralnick/millerjared/PhenoInterpolate/outputs/figures/response-curves-interp-status.png",
       width = 20, height = 12)
print(combined_response)

# --- Summary table ---
error_summary <- error_long %>%
  group_by(model, interp_status, year) %>%
  summarise(
    mean_abs_error   = round(mean(abs_error, na.rm = TRUE), 2),
    median_abs_error = round(median(abs_error, na.rm = TRUE), 2),
    n_cells = n(),
    .groups = "drop"
  )
print(error_summary)

plot(gam_spatial)
plot(gam_env)
```

Compare GAM to true model fit

```{r}
#| output: true
# --- Build prediction grid over observed predictor ranges ---
pred_grid <- expand.grid(
  tmean_spring_mean = seq(min(cells_true_flowering$tmean_spring, na.rm = TRUE),
                          max(cells_true_flowering$tmean_spring, na.rm = TRUE),
                          length.out = 100),
  ppt_spring_mean   = c(20, 30, 55, 95)  # same ppt scenarios as response curve viz
) %>%
  mutate(ppt_label = factor(ppt_spring_mean,
                             levels = c(20, 30, 55, 95),
                             labels = c("Drought (<25mm)", "Adequate (30mm)",
                                        "Optimal (55mm)", "Excess (>85mm)")))

# --- Get GAM predictions with SE ---
pred_grid$predicted_spatial <- predict(gam_spatial,
                                        newdata = pred_grid,
                                        exclude = "s(cluster_full_id)",
                                        newdata.guaranteed = TRUE,
                                        se.fit = FALSE)
pred_grid$predicted_env <- predict(gam_env,
                                    newdata = pred_grid,
                                    exclude = "s(cluster_full_id)",
                                    newdata.guaranteed = TRUE,
                                    se.fit = FALSE)

# --- Build true curve for same grid ---
pred_grid <- pred_grid %>%
  rowwise() %>%
  mutate(
    ppt_penalty = if(ppt_spring_mean < params$ppt_min){
      params$ppt_max_penalty
    } else if(ppt_spring_mean > params$ppt_max){
      min(((ppt_spring_mean - params$ppt_max) / params$ppt_max) * params$ppt_max_penalty,
          params$ppt_max_penalty)
    } else {
      0
    },
    true_onset = qnorm(0.10,
                       mean = params$intercept + params$temp_slope * tmean_spring_mean + ppt_penalty,
                       sd = params$sigma_individual)
  ) %>%
  ungroup()

# --- Pivot to long for faceted plotting ---
pred_long <- pred_grid %>%
  tidyr::pivot_longer(cols = c(predicted_spatial, predicted_env, true_onset),
               names_to = "source",
               values_to = "onset") %>%
  mutate(source = recode(source,
                         "predicted_spatial" = "GAM (Spatial)",
                         "predicted_env"     = "GAM (Environmental)",
                         "true_onset"        = "True Response"))

# --- Plot ---
gam_vs_true <- ggplot(pred_long, aes(x = tmean_spring_mean, y = onset,
                                      color = source, linetype = source)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("GAM (Spatial)"       = "steelblue",
                                 "GAM (Environmental)" = "darkorange",
                                 "True Response"            = "black")) +
  scale_linetype_manual(values = c("GAM (Spatial)"       = "dashed",
                                    "GAM (Environmental)" = "dashed",
                                    "True Response"            = "solid")) +
  facet_wrap(~ppt_label) +
  labs(x = "Mean Spring Temperature (°C)", y = "Flowering Onset (DOY)",
       title = "GAM Estimated vs True Response Curve",
       color = "Source", linetype = "Source") +
  theme_bw()

print(gam_vs_true)
```
