---
title: "bee-plant-baseline-overlaps"
author: "JT Miller"
format: html
toc: true
theme: journal
self-contained: true  
editor: visual
code-fold: true
---

# Bees & Peas, using macroecological modeling to compare distributions, phenology, and interaction space.

### Background

Global change is shifting species across axes of space & time (Fredston et al. 2025). With these shifts, species interactions and ecological structure of communities are likely to alter, modifying both the strength of these interactions and the biodiversity that shapes them. Forming hypotheses for potential interaction space and how they may vary across each taxon's range is important for establishing baselines to both model central biological tendencies of these ecological partners, as well as forming priorities for conservation of these ecologies. For plants and pollinators, three macroecological elements determine the interaction potential: spatial overlap, temporal overlap, and visitation likelihood. To construct hypotheses of potential interaction space over the bulk of a pollinator range, I combined spatial distribution modeling, interpolated phenology modeling, and interaction data sourced from GLOBI to evaluate how legume pollen emphasizing bees and host-plant taxa vary in overlaps over the each bees species range. 

### Methods 

Load Libraries 
```{r}
library(sf)
library(terra)
library(tidyverse)
library(data.table)
library(dplyr)
library(ggplot2)
library(patchwork)
source("/blue/guralnick/millerjared/PhenoInterpolate/code/draft/proxy-baseline-fxns.R")
```


#### 1. Spatial Overlap
We fitted enviromental niche models per taxa using best practices to construct each species distribution. We can then overlap these distributions with taxa of known interactions to construct a comparison of where interacting partners overlap. 
```{r}
# Bring In Species Distribution Models
## CRS setup
target_crs = "+proj=moll +lon_0=-87.38457 +datum=WGS84 +units=m +no_defs"

## Load in Bee Species Distribution Model 
a_holo_sdm <- terra::rast(paste0("/blue/soltis/millerjared/Legume-Specialist-Occupancy/outputs/sdms/bees/", "Anthophora_hololeuca", "_SDM_PA.tif"))
a_holo_sdm <- terra::project(a_holo_sdm, target_crs, method = "near")

## Load in Host-Plants Species Distribution Models
## Plant models 
plant_sp_old <- c("Psorothamnus arborescens","Psorothamnus emoryi","Psorothamnus fremontii", ## deal with some name changes.
"Psorothamnus polydenius","Psorothamnus schottii","Psorothamnus spinosus")
plant_sp_current <- c("Psorodendron arborescens","Psorothamnus emoryi", "Psorodendron fremontii",
"Psorothamnus polydenius", "Psorodendron schottii", "Psorodendron spinosum")
plant_PA_sdms <- list()
  for(i in 1:length(plant_sp_old)){
    if(file.exists(paste0("/blue/soltis/millerjared/Legume-Specialist-Occupancy/outputs/sdms/plants/", gsub(" ", "_", plant_sp_old[i]), "_SDM_PA.tif"))){
      plant_PA_sdms[[i]] <- terra::rast(paste0("/blue/soltis/millerjared/Legume-Specialist-Occupancy/outputs/sdms/plants/", gsub(" ", "_", plant_sp_old[i]), "_SDM_PA.tif"))
    }
  }
plant_PA_sdms <- lapply(plant_PA_sdms, function(r) {
  terra::project(r, a_holo_sdm, method = "near")
})
names(plant_PA_sdms) <- plant_sp_current

## stack plant rasters
plant_sdm_stack <- terra::rast(plant_PA_sdms)

# Overlap Taxa
## mask plants to bee presences, only where bee == 1 and plant can be 0 or 1 
plants_in_bee_range <- plant_sdm_stack * a_holo_sdm
## calc plant sp richness across bee range
plant_richness <- sum(plants_in_bee_range, na.rm = TRUE)
## create P/A by cell for plant availability 
plant_availability <- as.data.frame(plants_in_bee_range, xy = TRUE, na.rm = TRUE)

## Create a richness df by pixel and visualize across bee's range
richness_df <- as.data.frame(plant_richness, xy = TRUE)
colnames(richness_df)[3] <- "richness"
basemap <- rnaturalearth::ne_states(c("United States of America", "Mexico"), returnclass = "sf")
basemap <- basemap %>% filter(name %in% c("California", "Nevada", "Arizona", "Sonora", "Baja California", "Baja California Sur"))
basemap_reproj <- basemap %>% st_transform(st_crs(target_crs))
  
ggplot() +
  geom_tile(data = richness_df, aes(x = x, y = y, fill = richness)) +
  geom_sf(data = basemap_reproj, fill = NA, color = "black", linewidth = 0.5) +
  scale_fill_viridis_c(name = "Host-Plant Species\nRichness") +
  coord_sf(crs = st_crs(basemap_reproj)) +
  theme_bw() +
  labs(title = "Host-Plant Species Richness in Anthophora hololeuca's Range",
       x = "Longitude", y = "Latitude")

## Create a faceted figure to show which host-plants are where in the bee's range 
plants_df <- as.data.frame(plants_in_bee_range, xy = TRUE) %>%
  pivot_longer(cols = -c(x, y), names_to = "species", values_to = "presence") %>%
  filter(!is.na(presence))

ggplot() +
  geom_tile(plants_df, mapping = aes(x = x, y = y, fill = factor(presence))) +
  geom_sf(data = basemap, fill = NA, color = "black", linewidth = 0.5) +
  facet_wrap(~species, ncol = 2) +
  scale_fill_manual(values = c("0" = "grey80", "1" = "darkgreen"),
                    name = "Presence") +
  coord_sf(crs = st_crs(basemap_reproj)) +
  theme_bw() +
  labs(title = "Individual Plant Species Presence within Anthophora hololeuca's Range",
       x = "Longitude", y = "Latitude")
```

We can take this a bit further and create cross-wise comparisons to see what host-plants are available per modeled pixel. 
```{r}
## first restrict to pixels where both the bee and plant co-occur
bee_plant_overlap <- plant_sdm_stack * a_holo_sdm # where bee and plant == 1
overlap_df <- as.data.frame(bee_plant_overlap, xy = TRUE)
species_cols <- names(overlap_df)[-c(1,2)] # retrieve plant species names for comparison
## Keep only rows where a plant has at least 1 occurrence
keep_rows <- rowSums(overlap_df[, -c(1,2)] == 1, na.rm = TRUE) > 0
filtered_df <- overlap_df[keep_rows, ]

## calc n species per pixel 
filtered_df$n_species <- rowSums(filtered_df[, -c(1,2)] == 1, na.rm = TRUE)

## create combinations using apply 
filtered_df$combination <- apply(filtered_df[, species_cols], 1, function(row) {
  present <- species_cols[!is.na(row) & row == 1]
  present_abbrev <- gsub("^Psorothamnus |^Psorodendron ", "P. ", present)
  paste(present_abbrev, collapse = " & ")
})

## Select to simplify
combination_df <- filtered_df[, c("x", "y", "combination", "n_species")]

## Summarize
combo_summary <- as.data.frame(table(combination_df$combination, combination_df$n_species))
colnames(combo_summary) <- c("combination", "n_species", "count")
combo_summary <- combo_summary[combo_summary$count > 0, ]
combo_summary <- combo_summary[order(-combo_summary$count), ]

# Create visuals 
## get top combinations ordered by count
top_combos <- combo_summary$combination[1:10]

## get singles for areas where the bee only has access to a single host plant species
singles <- combo_summary %>% filter(n_species == 1)

## filter data
combination_df_filtered <- combination_df[combination_df$combination %in% top_combos, ]
combination_df_singles <- combination_df %>% filter(combination %in% singles$combination)

## convert to factor with levels ordered by count (descending)
combination_df_filtered$combination <- factor(
  combination_df_filtered$combination,
  levels = combo_summary$combination[combo_summary$combination %in% top_combos]
)

combination_df_singles$combination <- factor(
  combination_df_singles$combination,
  levels = singles$combination
)

## plot 
ggplot() +
  geom_tile(data = combination_df_filtered, aes(x = x, y = y, fill = combination)) +
  geom_sf(data = basemap_reproj, fill = NA, color = "black", linewidth = 0.1, alpha = 0.2) +
  scale_fill_brewer(palette = "Set3", name = "Plant Species\nCombination") +
  coord_sf() +
  theme_bw() +
  labs(title = "Top 10 Most Common Host Plant Species Combinations for A. hololeuca Presence Overlap",
       x = "Longitude", y = "Latitude")

ggsave("/blue/guralnick/millerjared/PhenoInterpolate/outputs/figures/top-ten-host-combos-for-a-holo-range.png", height = 10, width = 10)


ggplot() +
  geom_tile(data = combination_df_singles, aes(x = x, y = y, fill = combination)) +
  geom_sf(data = basemap_reproj, fill = NA, color = "black", linewidth = 0.1, alpha = 0.2) +
  scale_fill_brewer(palette = "Set3", name = "Plant Species\nCombination") +
  coord_sf() +
  theme_bw() +
  labs(title = "Regions where only a single host plant is present for A. hololeuca Presence Overlap",
       x = "Longitude", y = "Latitude")

ggsave("/blue/guralnick/millerjared/PhenoInterpolate/outputs/figures/single-host-for-a-holo-range.png", height = 10, width = 10)

ggplot() +
  geom_tile(combination_df, mapping = aes(x = x, y = y, fill = factor(n_species))) +
  geom_sf(data = basemap_reproj, fill = NA, color = "black", linewidth = 0.1, alpha = 0.2) +
  scale_fill_viridis_d(name = "Number of\nPlant Species") +
  coord_sf() +
  theme_bw() +
  labs(title = "Host Plant Species Richness in A. hololeuca's Range",
       x = "Longitude", y = "Latitude")

ggsave("/blue/guralnick/millerjared/PhenoInterpolate/outputs/figures/host-plant-richness-for-a-holo-range.png", height = 10, width = 10)
```

#### 2. Phenology Overlaps
Our goal here is to model predictions for onset and offset of flowering/insect flight per species across their range, so we may then compare pixel-by-pixel for potential overlaps. The challenge to this is that the empirical data that is available is opportunistic and messy. Here, I use Hopkins BioClimatic Law (Hopkins 1920) as a simple method for aggregating these data annually, constructing phenophase estimates using the package phenesse (Belitz et al. 2020), restructuring a latitude/elevation response based on empirical estimates, and finally using enviromental covariates to aid in the prediction of 'baseline' phenology trends across a taxon's range. <br>

Andrew Hopkins, a self taught entomologist (that received a honorary Doctorate...crazy) formalized a Bioclimatic Law that had been used historically to guide safe dates to sow wheat in order to avoid the damage of a Hessian Fly in Germany. He chose to use plants to formalize the law, recrding flowering and intensity across a temperate eastern gradient. His law suggests that for a taxon, there is a uniform response of delaying flowering (or emergence) by 4 days for every degree in latitude or 400 ft (122m) elevation (Hopkins 1920). While called a 'Law', its more like 'guidelines' when it comes to universal application. Hopkins noted this, specifically referring to the Rocky Mountains Plataeu as an area where this Law breaks, for reasons he described as *other* controlling influences not being equal across geography. I was simply interested in whether our data even followed such a pattern, and was pleasantly surprised by the results. While 4 day delays are not what we recover in the west, we *do* see strong signal in there being a predictable delay across increments of 1 degree of latitude or 122m in elevation (or a equation to achieve whatever combination of these two). So much so that we can fit a simple linear model to these binned data and see a linear response in central tendencies. 
```{r}
## Load BeeBDC data and Extra A. hololeuca data provided by M. Orr
bee_occs <- fread("/blue/soltis/millerjared/beeBDC/Bee-data/OutputData/05_cleaned_database.csv")
extra_hololeuca_data <- fread("/blue/soltis/millerjared/SpatioTemporalTradeoffs/data/raw/a-hololeuca-from-Orr.csv")

# Use a custom built function to grab target taxa's phenophase data (cleaned)
target_taxa <- "Anthophora hololeuca"
phen_data <- grab_bee_occs(target_taxa, bee_occs = bee_occs, extra_hololeuca_data = extra_hololeuca_data)
phen_data[, row_id := .I] # create a row index to run joins
# Add Elevation to Point Occurrence Data
## TIF of elevation data for western NA
na_elev <- rast("/blue/guralnick/millerjared/PhenoElevation/data/NAelevation4.tif")
## convert occ data to a spatVect in WGS84
phen_data_coords <- phen_data[!is.na(longitude) & !is.na(latitude)]
phen_data_vect <- vect(phen_data_coords, 
                         geom = c("longitude", "latitude"),
                         crs = "EPSG:4326")
## reproj to elevation raster projection
phen_data_vect_reproj1 <- project(phen_data_vect, crs(na_elev))
# Extract all raster values
elevation_vals <- terra::extract(na_elev, phen_data_vect_reproj1)
# append to coord data
phen_data_coords[, elevation_m := elevation_vals[,2]]
# merge with Orig datatable
phen_data <- phen_data[phen_data_coords[, .(row_id, elevation_m)],
                         on = "row_id"]
## Remove NA elev
phen_data <- phen_data %>% filter(!is.na(elevation_m))

# Create Hopkins Expectation: For every degree in latitude or 122 meters of elevation, flowering will be delayed by 4 days. 
  # use min vals as a baseline start
  min_lat <- min(phen_data$latitude)
  min_elev <- min(phen_data$elevation_m)
  ## create expected # of delay days based on hopkins
  phen_data[, hopkins_delay_days :=
              ((latitude - min_lat)* 4) + # latitudnal effect
              ((elevation_m - min_elev))/122 * 4] # elevation effect
  
  # Create bins for phenology data based on Hopkins delay by intervals of 4
  phen_data[, hopkins_bin := cut(hopkins_delay_days, 
                                 breaks = seq(0, max(hopkins_delay_days, na.rm = TRUE) + 4, by = 4), include.lowest = TRUE)]
  
# Before running phenometrics, its important to account for modality. This is best inspected visually, so I created a fxn that allows for assessments and to extract local minima in order to better limit where we make dayOfYear cutoff decisions
described_phen_months <- fread("/blue/soltis/millerjared/Legume-Specialist-Occupancy/data/raw/hololeuca-and-host-pheno.csv")
taxon_phen_hist(target_taxa, bees_or_plants = "bees", described_phen_months = described_phen_months)
min_doy <- 21
max_doy <- 234
## Run Phenometric Cleaning 
# remove outliers based on known flowering 
phen_data <- phen_data %>%  filter(dayOfYear > min_doy & dayOfYear < max_doy)
# downsample any dayOfYears that have abnormally high sampling on a dayOfYear within a year (City Nature Challenge and BioBlitzs)
phen_data <- phen_data %>% 
    group_by(species,year,hopkins_bin,dayOfYear) %>% 
    dplyr::slice_sample(n = 3)
# if sample is either a specimen (Herbarium data) or an occurrence (unspecified in bee data), then remove first day of the months from these data
# these are possibly a data quality issue
  phen_data <- phen_data %>% 
    filter(!(source %in% c("occurrence", "specimen") & day == 1))

   ## Create a phenesse estimate for phenometric 
  phen_data_est_onset <- phen_data %>% 
    group_by(species,hopkins_bin) %>% 
    filter(n() >= 7) %>%
    filter(n_distinct(dayOfYear) > 3) %>% 
    group_modify(~ broom::tidy(phenesse::quantile_ci(observations = .x$dayOfYear, percentile = 0.10, bootstraps=1000)))
  
  phen_data_est_offset <- phen_data %>% 
    group_by(species,hopkins_bin) %>% 
    filter(n() >= 7) %>%
    filter(n_distinct(dayOfYear) > 3) %>% 
    group_modify(~ broom::tidy(phenesse::quantile_ci(observations = .x$dayOfYear, percentile = 0.90, bootstraps=1000)))
  
  # extract estimate
  phen_onset_estimates <- phen_data_est_onset %>% 
    filter(column == "estimate") %>% 
    select(species, hopkins_bin, mean) %>% 
    rename(onset_doy = mean) %>% ungroup()
  
  phen_offset_estimates <- phen_data_est_offset %>% 
    filter(column == "estimate") %>% 
    select(species, hopkins_bin, mean) %>% 
    rename(offset_doy = mean) %>% ungroup() %>% select(-species)
  
  phen_estimates <- phen_onset_estimates %>% 
    left_join(phen_offset_estimates, by = "hopkins_bin")

# Remove any outliers 
  phen_estimates <-  phen_estimates %>% 
    group_by(species) %>% 
    mutate(zonset = scale(onset_doy)) %>% 
    filter(between(zonset,-3.25,+3.25))
  
  phen_estimates <-  phen_estimates %>% 
    group_by(species) %>% 
    mutate(zoffset = scale(offset_doy)) %>% 
    filter(between(zoffset,-3.25,+3.25))
  
## Use Species Anchoring To Create a Predicted Hopkins Onset
  # Use Rob's estimate sp anchors fxn
  HOPKINS_LAT_SLOPE <- 4
  HOPKINS_ELEV_SLOPE <- 4/122 
  estimate_species_anchors <- function(df) {
    df |> group_by(species) |>
      group_modify(~{
        x <- .x
        slat <- median(x$lat_mean, na.rm = TRUE)
        selev <- median(x$mid_elev, na.rm = TRUE)  
        
        if (nrow(x) >= 10 && length(unique(x$lat_mean[!is.na(x$lat_mean)])) > 2) {
          m <- lm(onset ~ lat_mean, data = x)
          sdoy <- as.numeric(predict(m, newdata = data.frame(lat_mean = slat)))
        } else {
          sdoy <- median(x$onset, na.rm = TRUE)
        }
        tibble(start_lat = slat, start_elev = selev, start_doy = sdoy) 
      }) |>
      ungroup()
  }
  
  # bin summary (for creating hopkins predictions)
  bin_summary <- phen_data %>%
    group_by(species, hopkins_bin) %>%
    summarise(
      hopkins_lat = median(latitude, na.rm = TRUE),
      hopkins_elev = median(elevation_m, na.rm = TRUE),
      hopkins_delay = median(hopkins_delay_days, na.rm = TRUE),
      n_obs = n(),
      .groups = 'drop'
    )
  
  # join bins 
  phen_estimates <- phen_estimates %>% 
    left_join(bin_summary, by = "hopkins_bin") %>% 
    rename(lat_mean = hopkins_lat, 
           mid_elev = hopkins_elev, 
           onset = onset_doy, 
           offset = offset_doy,
           species = species.x) # redundant
  
  anchored_sp <- estimate_species_anchors(phen_estimates)
  
  
  phen_estimate <- phen_estimates %>% 
    left_join(anchored_sp, by = 'species')
  
  # extract midpoint of each hopkins bin to use as predictor
  phen_estimate <- phen_estimate %>%
    mutate(
      # Extract numeric values from bin intervals
      bin_lower = as.numeric(sub("\\((.+),.+\\]", "\\1", hopkins_bin)),
      bin_upper = as.numeric(sub("\\(.+,(.+)\\]", "\\1", hopkins_bin)),
      hopkins_delay_midpoint = (bin_lower + bin_upper) / 2
    )
  
  ## Run a Hopkins model based on bins
  hopkins_model <- lm(onset ~ hopkins_delay_midpoint, data = phen_estimate)
  r2_hop <- performance::r2(hopkins_model)
  res_hop <- residuals(hopkins_model)
  rmse_val_hop <- sqrt(mean(res_hop^2))
  # Extract the corrected slope (actual delay rate per Hopkins bin unit)
  actual_delay_rate <- coef(hopkins_model)[2]
  actual_delay_se <- summary(hopkins_model)$coefficients[2, 2]
  # Apply the corrected slope with uncertainty bounds
  phen_estimate <- phen_estimate %>%
    mutate(
      #cCalculate delay from the anchor point using Hopkins bins
      delay_from_anchor = hopkins_delay_midpoint - 
        ((start_lat - min_lat) * 4 + 
           (start_elev - min_elev) / 122 * 4),
      
      # apply actual observed delay rate
      corrected_onset_prediction = start_doy + (delay_from_anchor * actual_delay_rate),
      
      # Add uncertainty bounds based on SE of slope
      corrected_onset_lower = start_doy + (delay_from_anchor * (actual_delay_rate - actual_delay_se)),
      corrected_onset_upper = start_doy + (delay_from_anchor * (actual_delay_rate + actual_delay_se))
    )

## Visualize 
hop_plot1 <- ggplot(phen_estimate, aes(x = hopkins_delay_midpoint, y = onset)) +
    geom_point(size = 3) +
    geom_smooth(method = "lm", se = TRUE) +
    annotate("text", x = min(phen_estimate$hopkins_delay_midpoint), y = max(phen_estimate$onset),
             label = sprintf("R² = %.3f\nRMSE = %.2f days\nActual Delay Rate = %.3f days", r2_hop[[2]], rmse_val_hop, actual_delay_rate),
             hjust = 0, vjust = 1) +
    labs(x = "Hopkins Expected Delay Bin (Middle Value for Bin)",
         y = "Observed Onset (day of year)",
         title = paste("Testing Hopkins Bioclimatic Law on", target_taxa[1])) +
    theme_minimal()

hop_plot2 <- ggplot(phen_estimate, aes(x = corrected_onset_prediction, y = onset)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
    #geom_ribbon(aes(ymin = corrected_onset_lower, ymax = corrected_onset_upper), 
    #alpha = 0.2, fill = "blue") +
    geom_point(size = 3) +
    annotate("text", x = min(phen_estimate$onset), y = max(phen_estimate$corrected_onset_prediction),
             label = sprintf("R² = %.3f\nRMSE = %.2f days", r2_hop[[2]], rmse_val_hop),
             hjust = 0, vjust = 1) +
    labs(x = "Corrected Predicted Onset (anchored + actual slope)",
         y = "Empirical Estimate Onset (day of year)",
         title = paste("Hopkins Estimate Error Plot", target_taxa[1])) +
    theme_minimal()

hop_plot1 + hop_plot2
  
```
As we can see, Hopkins Bins actually end up showing us a very convinient way to stratify the aggregation units for making phenometric estimates on the unstructured data. It is little suprise there is some utility to this (as latitude and elevation both modify the enviroment an organism experiences, principally by temperature), but to see such a strong signal particularly in a western desert taxa was not something I really expected. Whats really cool about this I think is that this relationship is strong enough that we can say something about the delay rate. According to these empirical estimates, A. hololeuca will have a delayed emergence of 1 day for every degree of latitude or 122 meters in elevation (or whatever combination of those leads to the bins). One important point to emphasize here however is this should be interpreted as the central tendency/baseline of the species. Average enviromental conditions across this latitudnal/elevational gradient are what likely contribute to the predictable pattern we're observing, therefore deviations away from that average (either microhabitat variation and/or intrannual variations) will cause deviations.  <br>

We can try this for the host-plants by wrapping this code up into a function
```{r}
phenovision_flowering_data <- fread("/blue/guralnick/millerjared/phenovision_data/phenovision_flowers_03_15_2026.csv")
cch2_annotated_specimens <- fread("/blue/soltis/millerjared/SpatioTemporalTradeoffs/data/processed/cch2-data/cch2-cleaned-and-aligned-data.csv")

check_hopkins_relationship <- function(target_taxa, phenovision_flowering_data = phenovision_flowering_data, cch2_annotated_specimens = cch2_annotated_specimens, min_doy, max_doy){
  phen_data <- grab_flowering_occs(target_taxa, phenovision_flowering_data = phenovision_flowering_data, cch2_annotated_specimens = cch2_annotated_specimens)
  phen_data[, row_id := .I] # create a row index to run joins
# Add Elevation to Point Occurrence Data
## TIF of elevation data for western NA
na_elev <- rast("/blue/guralnick/millerjared/PhenoElevation/data/NAelevation4.tif")
## convert occ data to a spatVect in WGS84
phen_data_coords <- phen_data[!is.na(longitude) & !is.na(latitude)]
phen_data_vect <- vect(phen_data_coords, 
                         geom = c("longitude", "latitude"),
                         crs = "EPSG:4326")
## reproj to elevation raster projection
phen_data_vect_reproj1 <- project(phen_data_vect, crs(na_elev))
# Extract all raster values
elevation_vals <- terra::extract(na_elev, phen_data_vect_reproj1)
# append to coord data
phen_data_coords[, elevation_m := elevation_vals[,2]]
# merge with Orig datatable
phen_data <- phen_data[phen_data_coords[, .(row_id, elevation_m)],
                         on = "row_id"]
## Remove NA elev
phen_data <- phen_data %>% filter(!is.na(elevation_m))

# Create Hopkins Expectation: For every degree in latitude or 122 meters of elevation, flowering will be delayed by 4 days. 
  # use min vals as a baseline start
  min_lat <- min(phen_data$latitude)
  min_elev <- min(phen_data$elevation_m)
  ## create expected # of delay days based on hopkins
  phen_data[, hopkins_delay_days :=
              ((latitude - min_lat)* 4) + # latitudnal effect
              ((elevation_m - min_elev))/122 * 4] # elevation effect
  
  # Create bins for phenology data based on Hopkins delay by intervals of 4
  phen_data[, hopkins_bin := cut(hopkins_delay_days, 
                                 breaks = seq(0, max(hopkins_delay_days, na.rm = TRUE) + 4, by = 4), include.lowest = TRUE)]
  
## Run Phenometric Cleaning 
# remove outliers based on known flowering 
phen_data <- phen_data %>%  filter(dayOfYear > min_doy & dayOfYear < max_doy)
# downsample any dayOfYears that have abnormally high sampling on a dayOfYear within a year (City Nature Challenge and BioBlitzs)
phen_data <- phen_data %>% 
    group_by(species,year,hopkins_bin,dayOfYear) %>% 
    dplyr::slice_sample(n = 3)
# if sample is either a specimen (Herbarium data) or an occurrence (unspecified in bee data), then remove first day of the months from these data
# these are possibly a data quality issue
  phen_data <- phen_data %>% 
    filter(!(source %in% c("occurrence", "specimen") & day == 1))

   ## Create a phenesse estimate for phenometric 
  phen_data_est_onset <- phen_data %>% 
    group_by(species,hopkins_bin) %>% 
    filter(n() >= 7) %>%
    filter(n_distinct(dayOfYear) > 3) %>% 
    group_modify(~ broom::tidy(phenesse::quantile_ci(observations = .x$dayOfYear, percentile = 0.10, bootstraps=1000)))
  
  phen_data_est_offset <- phen_data %>% 
    group_by(species,hopkins_bin) %>% 
    filter(n() >= 7) %>%
    filter(n_distinct(dayOfYear) > 3) %>% 
    group_modify(~ broom::tidy(phenesse::quantile_ci(observations = .x$dayOfYear, percentile = 0.90, bootstraps=1000)))
  
  # extract estimate
  phen_onset_estimates <- phen_data_est_onset %>% 
    filter(column == "estimate") %>% 
    select(species, hopkins_bin, mean) %>% 
    rename(onset_doy = mean) %>% ungroup()
  
  phen_offset_estimates <- phen_data_est_offset %>% 
    filter(column == "estimate") %>% 
    select(species, hopkins_bin, mean) %>% 
    rename(offset_doy = mean) %>% ungroup() %>% select(-species)
  
  phen_estimates <- phen_onset_estimates %>% 
    left_join(phen_offset_estimates, by = "hopkins_bin")

# Remove any outliers 
  phen_estimates <-  phen_estimates %>% 
    group_by(species) %>% 
    mutate(zonset = scale(onset_doy)) %>% 
    filter(between(zonset,-3.25,+3.25))
  
  phen_estimates <-  phen_estimates %>% 
    group_by(species) %>% 
    mutate(zoffset = scale(offset_doy)) %>% 
    filter(between(zoffset,-3.25,+3.25))
  
## Use Species Anchoring To Create a Predicted Hopkins Onset
  # Use Rob's estimate sp anchors fxn
  HOPKINS_LAT_SLOPE <- 4
  HOPKINS_ELEV_SLOPE <- 4/122 
  estimate_species_anchors <- function(df) {
    df |> group_by(species) |>
      group_modify(~{
        x <- .x
        slat <- median(x$lat_mean, na.rm = TRUE)
        selev <- median(x$mid_elev, na.rm = TRUE)  
        
        if (nrow(x) >= 10 && length(unique(x$lat_mean[!is.na(x$lat_mean)])) > 2) {
          m <- lm(onset ~ lat_mean, data = x)
          sdoy <- as.numeric(predict(m, newdata = data.frame(lat_mean = slat)))
        } else {
          sdoy <- median(x$onset, na.rm = TRUE)
        }
        tibble(start_lat = slat, start_elev = selev, start_doy = sdoy) 
      }) |>
      ungroup()
  }
  
  # bin summary (for creating hopkins predictions)
  bin_summary <- phen_data %>%
    group_by(species, hopkins_bin) %>%
    summarise(
      hopkins_lat = median(latitude, na.rm = TRUE),
      hopkins_elev = median(elevation_m, na.rm = TRUE),
      hopkins_delay = median(hopkins_delay_days, na.rm = TRUE),
      n_obs = n(),
      .groups = 'drop'
    )
  
  # join bins 
  phen_estimates <- phen_estimates %>% 
    left_join(bin_summary, by = "hopkins_bin") %>% 
    rename(lat_mean = hopkins_lat, 
           mid_elev = hopkins_elev, 
           onset = onset_doy, 
           offset = offset_doy,
           species = species.x) # redundant
  
  anchored_sp <- estimate_species_anchors(phen_estimates)
  
  
  phen_estimate <- phen_estimates %>% 
    left_join(anchored_sp, by = 'species')
  
  # extract midpoint of each hopkins bin to use as predictor
  phen_estimate <- phen_estimate %>%
    mutate(
      # Extract numeric values from bin intervals
      bin_lower = as.numeric(sub("\\((.+),.+\\]", "\\1", hopkins_bin)),
      bin_upper = as.numeric(sub("\\(.+,(.+)\\]", "\\1", hopkins_bin)),
      hopkins_delay_midpoint = (bin_lower + bin_upper) / 2
    )
  
  ## Run a Hopkins model based on bins
  hopkins_model <- lm(onset ~ hopkins_delay_midpoint, data = phen_estimate)
  r2_hop <- performance::r2(hopkins_model)
  res_hop <- residuals(hopkins_model)
  rmse_val_hop <- sqrt(mean(res_hop^2))
  # Extract the corrected slope (actual delay rate per Hopkins bin unit)
  actual_delay_rate <- coef(hopkins_model)[2]
  actual_delay_se <- summary(hopkins_model)$coefficients[2, 2]
  # Apply the corrected slope with uncertainty bounds
  phen_estimate <- phen_estimate %>%
    mutate(
      #cCalculate delay from the anchor point using Hopkins bins
      delay_from_anchor = hopkins_delay_midpoint - 
        ((start_lat - min_lat) * 4 + 
           (start_elev - min_elev) / 122 * 4),
      
      # apply actual observed delay rate
      corrected_onset_prediction = start_doy + (delay_from_anchor * actual_delay_rate),
      
      # Add uncertainty bounds based on SE of slope
      corrected_onset_lower = start_doy + (delay_from_anchor * (actual_delay_rate - actual_delay_se)),
      corrected_onset_upper = start_doy + (delay_from_anchor * (actual_delay_rate + actual_delay_se))
    )

## Visualize 
hop_plot1 <- ggplot(phen_estimate, aes(x = hopkins_delay_midpoint, y = onset)) +
    geom_point(size = 3) +
    geom_smooth(method = "lm", se = TRUE) +
    annotate("text", x = min(phen_estimate$hopkins_delay_midpoint), y = max(phen_estimate$onset),
             label = sprintf("R² = %.3f\nRMSE = %.2f days\nActual Delay Rate = %.3f days", r2_hop[[2]], rmse_val_hop, actual_delay_rate),
             hjust = 0, vjust = 1) +
    labs(x = "Hopkins Expected Delay Bin (Middle Value for Bin)",
         y = "Observed Onset (day of year)",
         title = paste("Testing Hopkins Bioclimatic Law on", target_taxa[1])) +
    theme_minimal()

hop_plot2 <- ggplot(phen_estimate, aes(x = corrected_onset_prediction, y = onset)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
    #geom_ribbon(aes(ymin = corrected_onset_lower, ymax = corrected_onset_upper), 
    #alpha = 0.2, fill = "blue") +
    geom_point(size = 3) +
    annotate("text", x = min(phen_estimate$onset), y = max(phen_estimate$corrected_onset_prediction),
             label = sprintf("R² = %.3f\nRMSE = %.2f days", r2_hop[[2]], rmse_val_hop),
             hjust = 0, vjust = 1) +
    labs(x = "Corrected Predicted Onset (anchored + actual slope)",
         y = "Empirical Estimate Onset (day of year)",
         title = paste("Hopkins Estimate Error Plot", target_taxa[1])) +
    theme_minimal()

return(hop_plot1 + hop_plot2)
}

## Load in North American Taxonomy
na_taxonomy <- fread("/blue/guralnick/millerjared/PlantSweepeR/data/processed/wcvp-ncbi-alignment-na.csv")


### For each plant taxa, run this fxn 

target_taxa <- na_taxonomy %>% 
  filter(alignedParentName == "Psorodendron arborescens") %>% 
  arrange(desc(nameStatus == "Accepted"), desc(alignedParentName == name)) %>% 
  pull(name)

taxon_phen_hist(target_taxa, bees_or_plants = "plants", described_phen_months = described_phen_months)

check_hopkins_relationship(target_taxa, phenovision_flowering_data = phenovision_flowering_data, cch2_annotated_specimens = cch2_annotated_specimens, min_doy = 30, max_doy = 228)


target_taxa <- na_taxonomy %>% 
  filter(alignedParentName == "Psorothamnus polydenius") %>% 
  arrange(desc(nameStatus == "Accepted"), desc(alignedParentName == name)) %>% 
  pull(name)

taxon_phen_hist(target_taxa, bees_or_plants = "plants", described_phen_months = described_phen_months)

check_hopkins_relationship(target_taxa, phenovision_flowering_data = phenovision_flowering_data, cch2_annotated_specimens = cch2_annotated_specimens, min_doy = 38, max_doy = 244)



target_taxa <- na_taxonomy %>% 
  filter(alignedParentName == "Psorodendron fremontii") %>% 
  arrange(desc(nameStatus == "Accepted"), desc(alignedParentName == name)) %>% 
  pull(name)

taxon_phen_hist(target_taxa, bees_or_plants = "plants", described_phen_months = described_phen_months)

check_hopkins_relationship(target_taxa, phenovision_flowering_data = phenovision_flowering_data, cch2_annotated_specimens = cch2_annotated_specimens, min_doy = 19, max_doy = 238)



target_taxa <- na_taxonomy %>% 
  filter(alignedParentName == "Psorothamnus emoryi") %>% 
  arrange(desc(nameStatus == "Accepted"), desc(alignedParentName == name)) %>% 
  pull(name)

taxon_phen_hist(target_taxa, bees_or_plants = "plants", described_phen_months = described_phen_months)

check_hopkins_relationship(target_taxa, phenovision_flowering_data = phenovision_flowering_data, cch2_annotated_specimens = cch2_annotated_specimens, min_doy = 12, max_doy = 185)


target_taxa <- na_taxonomy %>% 
  filter(alignedParentName == "Psorodendron schottii") %>% 
  arrange(desc(nameStatus == "Accepted"), desc(alignedParentName == name)) %>% 
  pull(name)

taxon_phen_hist(target_taxa, bees_or_plants = "plants", described_phen_months = described_phen_months)

check_hopkins_relationship(target_taxa, phenovision_flowering_data = phenovision_flowering_data, cch2_annotated_specimens = cch2_annotated_specimens, min_doy = 15, max_doy = 215)

target_taxa <- na_taxonomy %>% 
  filter(alignedParentName == "Psorodendron spinosum") %>% 
  arrange(desc(nameStatus == "Accepted"), desc(alignedParentName == name)) %>% 
  pull(name)

taxon_phen_hist(target_taxa, bees_or_plants = "plants", described_phen_months = described_phen_months)

check_hopkins_relationship(target_taxa, phenovision_flowering_data = phenovision_flowering_data, cch2_annotated_specimens = cch2_annotated_specimens, min_doy = 86, max_doy = 248)

## Annual Taxa 
target_taxa <- na_taxonomy %>% 
  filter(alignedParentName == "Rafinesquia neomexicana") %>% 
  arrange(desc(nameStatus == "Accepted"), desc(alignedParentName == name)) %>% 
  pull(name)

taxon_phen_hist(target_taxa, bees_or_plants = "plants", described_phen_months = described_phen_months)

check_hopkins_relationship(target_taxa, phenovision_flowering_data = phenovision_flowering_data, cch2_annotated_specimens = cch2_annotated_specimens, min_doy = 0, max_doy = 217)



```

To me, these results are pretty good. We can see at least some relationship in most of the species with Hopkins BioClimatic Law. It works just as well for a annual plant (Rafinesquia neomexicana) as it does for perennials. <br> 

Lets take it a step further though, we're going to keep the binning to hopkins logic, but rather than using the estimated delay by bin as a predictor, we're going to use summaries of the average climatic conditions for the phenophase observed per bin to fit a generalized linear model. Per flowering/bee occurrence, I extracted bioclimatic variables underlying that spatial point. For each hopkins delay bin, we then take the average of these occurrence's enviromental conditions. To model phenological onset as a function of climate, I then used an ensemble modeling framework with multicollinearity constraints and coefficient stability checks. For a pool of 19 annual bioclimatic variables (Chelsea v2.1) I generated all 3-predictor combinations (969 combos). All predictors were standardized with a z-score prior to modeling so that we could compare coefficient estimates across model fits. <br>

For each 3-predictor combination, we fit a generalized linear model with phenological onset as the response variable. I used performance::check_collinearity() to evaluate variance inflation factors (VIF) and only retained models with a VIF <= 5 for all predictors. From the valid models, I ran model selection, and retained the top 10 models (AIC < 2) if applicable (as these models are competitive for explanatory power), or otherwise just the top model (if AIC > 2 for any singular model). <br>

If multiple models were identified (AIC < 2), I assessed coefficient stability across the ensemble candidates. For each predictor appearing in the candidate models, we calculated the coefficient of variation (CV) and relative interquantile range (IQR) across models. If any predictor had high variability (CV > 50% or IQR > 0.5), this indcated that the predictors effects were unstable across models, and therefore I chose to abandom ensemble methods and instead selected the single best model based on AIC. If coefficient stability was found, I proceeded with ensembling based on normalized AIC weights. 


```{r}
model_baseline_onset_w_clim <- function(target_taxa,
                                     bees_or_plants = "plants", # target taxon group to pull empirical data for
                                     phenovision_flowering_data, # input inat flowering annotations for flowering plants
                                     cch2_annotated_specimens,  # input specimen flowering annotations for flowering plants (CA)
                                     bee_occs, # input bee occurrence data
                                     extra_hololeuca_data, # extra A. hololeuca occurrence data provided by M. Orr
                                     min_doy = 0, # look at histogram when making this decision
                                     max_doy = 365 # ditto to above
){
  
  ## Get Phenology Data
  if(bees_or_plants == "plants"){
    phen_data <- grab_flowering_occs(target_taxa, phenovision_flowering_data = phenovision_flowering_data, cch2_annotated_specimens = cch2_annotated_specimens)
  }
  if(bees_or_plants == "bees"){
    phen_data <- grab_bee_occs(target_taxa, bee_occs = bee_occs, extra_hololeuca_data = extra_hololeuca_data)
  }
  
  ## Attach Elevation & Climatic Variable Data
  # add a row ID to track where we're combing vars with occs
  phen_data[, row_id := .I]
  phen_data_coords <- phen_data[!is.na(longitude) & !is.na(latitude)]
  # load North America Elev Raster & Chelsa annual clim vars
  na_elev <- rast("/blue/guralnick/millerjared/PhenoElevation/data/NAelevation4.tif")
  avg_temp <- rast("/blue/guralnick/share/Chelsa-climate-data/climatologies/CHELSA_bio1_1981-2010_V.2.1.tif")
  avg_diurnal_temp_range <- rast("/blue/guralnick/share/Chelsa-climate-data/climatologies/CHELSA_bio02_1981-2010_V.2.1.tif")
  isothermality <- rast("/blue/guralnick/share/Chelsa-climate-data/climatologies/CHELSA_bio03_1981-2010_V.2.1.tif")
  temp_seas <- rast("/blue/guralnick/share/Chelsa-climate-data/climatologies/CHELSA_bio4_1981-2010_V.2.1.tif")
  tmax <- rast("/blue/guralnick/share/Chelsa-climate-data/climatologies/CHELSA_bio5_1981-2010_V.2.1.tif")
  tmin <- rast("/blue/guralnick/share/Chelsa-climate-data/climatologies/CHELSA_bio6_1981-2010_V.2.1.tif")
  annual_temp_range <- rast("/blue/guralnick/share/Chelsa-climate-data/climatologies/CHELSA_bio07_1981-2010_V.2.1.tif")
  avg_monthly_temp_wettest_q <- rast("/blue/guralnick/share/Chelsa-climate-data/climatologies/CHELSA_bio08_1981-2010_V.2.1.tif")
  avg_monthly_temp_driest_q <- rast("/blue/guralnick/share/Chelsa-climate-data/climatologies/CHELSA_bio09_1981-2010_V.2.1.tif")
  avg_monthly_temp_warmest_q <- rast("/blue/guralnick/share/Chelsa-climate-data/climatologies/CHELSA_bio10_1981-2010_V.2.1.tif")
  avg_monthly_temp_coldest_q <- rast("/blue/guralnick/share/Chelsa-climate-data/climatologies/CHELSA_bio11_1981-2010_V.2.1.tif")
  annual_precip <- rast("/blue/guralnick/share/Chelsa-climate-data/climatologies/CHELSA_bio12_1981-2010_V.2.1.tif")
  precip_wettest_month <- rast("/blue/guralnick/share/Chelsa-climate-data/climatologies/CHELSA_bio13_1981-2010_V.2.1.tif")
  precip_driest_month <- rast("/blue/guralnick/share/Chelsa-climate-data/climatologies/CHELSA_bio14_1981-2010_V.2.1.tif")
  precip_seas <- rast("/blue/guralnick/share/Chelsa-climate-data/climatologies/CHELSA_bio15_1981-2010_V.2.1.tif")
  avg_monthly_precip_wettest_q <- rast("/blue/guralnick/share/Chelsa-climate-data/climatologies/CHELSA_bio16_1981-2010_V.2.1.tif")
  avg_monthly_precip_driest_q <- rast("/blue/guralnick/share/Chelsa-climate-data/climatologies/CHELSA_bio17_1981-2010_V.2.1.tif")
  avg_monthly_precip_warmest_q <- rast("/blue/guralnick/share/Chelsa-climate-data/climatologies/CHELSA_bio18_1981-2010_V.2.1.tif")
  avg_monthly_precip_coldest_q <- rast("/blue/guralnick/share/Chelsa-climate-data/climatologies/CHELSA_bio19_1981-2010_V.2.1.tif")
  
  # convert occ data to a spatVect in WGS84
  phen_data_vect <- vect(phen_data_coords, 
                         geom = c("longitude", "latitude"),
                         crs = "EPSG:4326")
  # reproj to elevation raster projection & clim raster projections
  phen_data_vect_reproj1 <- project(phen_data_vect, crs(na_elev))
  phen_data_vect_reproj2 <- project(phen_data_vect, crs(avg_temp)) 
  # Stack all climate layers
  climate_stack <- c(avg_temp, avg_diurnal_temp_range, isothermality,
                     temp_seas, tmax, tmin, annual_temp_range,
                     avg_monthly_temp_wettest_q, avg_monthly_temp_driest_q,
                     avg_monthly_temp_warmest_q, avg_monthly_temp_coldest_q,
                     annual_precip, precip_wettest_month, precip_driest_month,
                     precip_seas, avg_monthly_precip_wettest_q, avg_monthly_precip_driest_q,
                     avg_monthly_precip_warmest_q, avg_monthly_precip_coldest_q)
  
  # make names clearer
  names(climate_stack) <- c("avg_temp", "avg_diurnal_temp_range", "isothermality",
                            "temp_seas", "tmax", "tmin", "annual_temp_range",
                            "avg_monthly_temp_wettest_q", "avg_monthly_temp_driest_q",
                            "avg_monthly_temp_warmest_q", "avg_monthly_temp_coldest_q",
                            "annual_precip", "precip_wettest_month", "precip_driest_month",
                            "precip_seas", "avg_monthly_precip_wettest_q", "avg_monthly_precip_driest_q",
                            "avg_monthly_precip_warmest_q", "avg_monthly_precip_coldest_q")
  

  # Extract all raster values
  elevation_vals <- terra::extract(na_elev, phen_data_vect_reproj1)
  climate_vals <- terra::extract(climate_stack, phen_data_vect_reproj2)
  
  # Add all values to these data
  phen_data_coords[, elevation_m := elevation_vals[,2]]
  phen_data_coords[, names(climate_vals)[-1] := climate_vals[,-1]]

  # add elevation values to occ table using row_id as our key
  phen_data_coords[, elevation_m := elevation_vals[,2]]
  # merge
  phen_data <- phen_data[phen_data_coords[, .(row_id, elevation_m, avg_temp, avg_diurnal_temp_range, isothermality,
                                              temp_seas, tmax, tmin, annual_temp_range,
                                              avg_monthly_temp_wettest_q, avg_monthly_temp_driest_q,
                                              avg_monthly_temp_warmest_q, avg_monthly_temp_coldest_q,
                                              annual_precip, precip_wettest_month, precip_driest_month,
                                              precip_seas, avg_monthly_precip_wettest_q, avg_monthly_precip_driest_q,
                                              avg_monthly_precip_warmest_q, avg_monthly_precip_coldest_q)],
                         on = "row_id"]
  # remove NA elevation vals
  phen_data <- phen_data[!is.na(elevation_m) & !is.na(avg_temp) & !is.na(avg_diurnal_temp_range) & !is.na(isothermality) & !is.na(temp_seas) 
                         & !is.na(tmax) & !is.na(tmin) & !is.na(annual_temp_range) & !is.na(avg_monthly_temp_wettest_q) & !is.na(avg_monthly_temp_driest_q) &
                           !is.na(avg_monthly_temp_warmest_q) & !is.na(avg_monthly_temp_coldest_q) & !is.na(annual_precip) &
                           !is.na(precip_wettest_month) & !is.na(precip_driest_month) & !is.na(precip_seas) & !is.na(avg_monthly_precip_wettest_q) &
                           !is.na(avg_monthly_precip_driest_q) & !is.na(avg_monthly_precip_warmest_q) & !is.na(avg_monthly_precip_coldest_q),]
  
  ## Create Hopkins Expectation: For every degree in latitude or 120 meters of elevation, flowering will be delayed by 4 days. 
  # use min vals as a baseline start
  min_lat <- min(phen_data$latitude)
  min_elev <- min(phen_data$elevation_m)
  ## create expected # of delay days based on hopkins
  phen_data[, hopkins_delay_days :=
              ((latitude - min_lat)* 4) + # latitudnal effect
              ((elevation_m - min_elev))/122 * 4] # elevation effect
  
  # Create bins for phenology data based on Hopkins delay by intervals of 4
  phen_data[, hopkins_bin := cut(hopkins_delay_days, 
                                 breaks = seq(0, max(hopkins_delay_days, na.rm = TRUE) + 4, by = 4), include.lowest = TRUE)]
  
  ## Run Phenometric Cleaning 
  # remove outliers based on known flowering 
  phen_data <- phen_data %>%  filter(dayOfYear > min_doy & dayOfYear < max_doy)
  # downsample any dayOfYears that have abnormally high sampling on a dayOfYear within a year (City Nature Challenge and BioBlitzs)
  phen_data <- phen_data %>% 
    group_by(species,year,hopkins_bin,dayOfYear) %>% 
    dplyr::slice_sample(n = 3)
  # if sample is either a specimen (Herbarium data) or an occurrence (unspecified in bee data), then remove first day of the months from these data
  # these are possibly a data quality issue
  phen_data <- phen_data %>% 
    filter(!(source %in% c("occurrence", "specimen") & day == 1))
  
  
  ## Create a phenesse estimate for phenometric 
  phen_data_est_onset <- phen_data %>% 
    group_by(species,hopkins_bin) %>% 
    filter(n() >= 7) %>%
    filter(n_distinct(dayOfYear) > 3) %>% 
    group_modify(~ broom::tidy(phenesse::quantile_ci(observations = .x$dayOfYear, percentile = 0.10, bootstraps=1000)))
  
  
  # extract estimate
  phen_onset_estimates <- phen_data_est_onset %>% 
    filter(column == "estimate") %>% 
    select(species, hopkins_bin, mean) %>% 
    rename(onset_doy = mean) %>% ungroup()
  
  phen_estimates <- phen_onset_estimates 
  

  
  ## Attach env avg summaries per hopkins bin
  hopkins_bin_env_summaries <- phen_data %>%
    group_by(species, hopkins_bin) %>%
    filter(n() >= 7) %>%
    filter(n_distinct(dayOfYear) > 3) %>%
    summarize(avg_temp = mean(avg_temp),
              avg_diurnal_temp_range = mean(avg_diurnal_temp_range),
              isothermality = mean(isothermality),
              temp_seas = mean(temp_seas),
              tmax = mean(tmax),
              tmin = mean(tmin),
              annual_temp_range = mean(annual_temp_range),
              avg_monthly_temp_wettest_q = mean(avg_monthly_temp_wettest_q),
              avg_monthly_temp_driest_q = mean(avg_monthly_temp_driest_q),
              avg_monthly_temp_warmest_q = mean(avg_monthly_temp_warmest_q),
              avg_monthly_temp_coldest_q = mean(avg_monthly_temp_coldest_q),
              annual_precip = mean(annual_precip),
              precip_wettest_month = mean(precip_wettest_month),
              precip_driest_month = mean(precip_driest_month),
              precip_seas = mean(precip_seas),
              avg_monthly_precip_wettest_q = mean(avg_monthly_precip_wettest_q),
              avg_monthly_precip_driest_q = mean(avg_monthly_precip_driest_q),
              avg_monthly_precip_warmest_q = mean(avg_monthly_precip_warmest_q),
              avg_monthly_precip_coldest_q = mean(avg_monthly_precip_coldest_q)
              
    ) %>%
    ungroup() %>%
    dplyr::select(hopkins_bin, avg_temp, avg_diurnal_temp_range, isothermality,
                  temp_seas, tmax, tmin, annual_temp_range, avg_monthly_temp_wettest_q,
                  avg_monthly_temp_driest_q, avg_monthly_temp_warmest_q, avg_monthly_temp_coldest_q,
                  annual_precip, precip_wettest_month, precip_driest_month, precip_seas,
                  avg_monthly_precip_wettest_q, avg_monthly_precip_driest_q, avg_monthly_precip_warmest_q,
                  avg_monthly_precip_coldest_q)
  ## Join 
  phen_estimates <- phen_estimates %>% 
    inner_join(hopkins_bin_env_summaries, by = "hopkins_bin")
  
  ## Remove Outliers 
  phen_estimates <-  phen_estimates %>% 
    group_by(species) %>% 
    mutate(zonset = scale(onset_doy)) %>% 
    filter(between(zonset,-3.25,+3.25))
  
  
  ## Use Species Anchoring To Create a Predicted Hopkins Onset
  # Use Rob's estimate sp anchors fxn
  HOPKINS_LAT_SLOPE <- 4
  HOPKINS_ELEV_SLOPE <- 4/122
  estimate_species_anchors <- function(df) {
    df |> group_by(species) |>
      group_modify(~{
        x <- .x
        slat <- median(x$lat_mean, na.rm = TRUE)
        selev <- median(x$mid_elev, na.rm = TRUE)  
        
        if (nrow(x) >= 10 && length(unique(x$lat_mean[!is.na(x$lat_mean)])) > 2) {
          m <- lm(onset ~ lat_mean, data = x)
          sdoy <- as.numeric(predict(m, newdata = data.frame(lat_mean = slat)))
        } else {
          sdoy <- median(x$onset, na.rm = TRUE)
        }
        tibble(start_lat = slat, start_elev = selev, start_doy = sdoy) 
      }) |>
      ungroup()
  }
  
  # bin summary (for creating hopkins predictions)
  bin_summary <- phen_data %>%
    group_by(species, hopkins_bin) %>%
    summarise(
      hopkins_lat = median(latitude, na.rm = TRUE),
      hopkins_elev = median(elevation_m, na.rm = TRUE),
      hopkins_delay = median(hopkins_delay_days, na.rm = TRUE),
      n_obs = n(),
      .groups = 'drop'
    )
  
  # join bins 
  phen_estimates <- phen_estimates %>% 
    left_join(bin_summary, by = "hopkins_bin") %>% 
    rename(lat_mean = hopkins_lat, 
           mid_elev = hopkins_elev, 
           onset = onset_doy, 
           species = species.x) # redundant
  
  anchored_sp <- estimate_species_anchors(phen_estimates)
  
  
  phen_estimate <- phen_estimates %>% 
    left_join(anchored_sp, by = 'species')
  
  # extract midpoint of each hopkins bin to use as predictor
  phen_estimate <- phen_estimate %>%
    mutate(
      # Extract numeric values from bin intervals
      bin_lower = as.numeric(sub("\\((.+),.+\\]", "\\1", hopkins_bin)),
      bin_upper = as.numeric(sub("\\(.+,(.+)\\]", "\\1", hopkins_bin)),
      hopkins_delay_midpoint = (bin_lower + bin_upper) / 2
    )
  
  ## Run a Hopkins model based on bins
  hopkins_model <- lm(onset ~ hopkins_delay_midpoint, data = phen_estimate)
  r2_hop <- performance::r2(hopkins_model)
  res_hop <- residuals(hopkins_model)
  rmse_val_hop <- sqrt(mean(res_hop^2))
  # Extract the corrected slope (actual delay rate per Hopkins bin unit)
  actual_delay_rate <- coef(hopkins_model)[2]
  actual_delay_se <- summary(hopkins_model)$coefficients[2, 2]
  # Apply the corrected slope with uncertainty bounds
  phen_estimate <- phen_estimate %>%
    mutate(
      #cCalculate delay from the anchor point using Hopkins bins
      delay_from_anchor = hopkins_delay_midpoint - 
        ((start_lat - min_lat) * 4 + 
           (start_elev - min_elev) / 120 * 4),
      
      # apply actual observed delay rate
      corrected_onset_prediction = start_doy + (delay_from_anchor * actual_delay_rate),
      
      # Add uncertainty bounds based on SE of slope
      corrected_onset_lower = start_doy + (delay_from_anchor * (actual_delay_rate - actual_delay_se)),
      corrected_onset_upper = start_doy + (delay_from_anchor * (actual_delay_rate + actual_delay_se))
    )
  
  #### ONSET ENV MODEL ######################################################################################
  if(nrow(phen_estimate) > 3){ # min req to make env
    predictors <- c("avg_temp", "avg_diurnal_temp_range", "isothermality",
                    "temp_seas", "tmax", "tmin", "annual_temp_range",
                    "avg_monthly_temp_wettest_q", "avg_monthly_temp_driest_q",
                    "avg_monthly_temp_warmest_q", "avg_monthly_temp_coldest_q",
                    "annual_precip", "precip_wettest_month", "precip_driest_month",
                    "precip_seas", "avg_monthly_precip_wettest_q", "avg_monthly_precip_driest_q",
                    "avg_monthly_precip_warmest_q", "avg_monthly_precip_coldest_q") 
    
    ## scale predictors using scale()
    phen_estimate_scaled <- phen_estimate
    scaling_params <- list()
    
    for(pred in predictors) {
      if(pred %in% names(phen_estimate)) {
        scaled_var <- scale(phen_estimate[[pred]])
        phen_estimate_scaled[[pred]] <- as.vector(scaled_var)
        scaling_params[[pred]] <- list(
          center = attr(scaled_var, "scaled:center"),
          scale = attr(scaled_var, "scaled:scale")
        )
      }
    }
    
    ## generate all combinations of 3 predictors
    all_combos <- combn(predictors, 3, simplify = FALSE)
    
    ## fit models with model selection and VIF checks
    vif_threshold <- 5
    model_results <- list()
    
    for(pred_combo in all_combos) {
      # Start with 3-predictor model (using scaled data)
      formula_str <- paste("onset ~", paste(pred_combo, collapse = " + "))
      full_model <- glm(as.formula(formula_str), data = phen_estimate_scaled)
      
      ## apply stepwise selection
      best_model <- MASS::stepAIC(full_model, direction = "both", trace = FALSE)
      
      ## count number of predictors
      n_predictors <- length(coef(best_model)) - 1
      
      ## check VIF only if 2+ predictors
      passes_vif <- TRUE
      max_vif <- NA
      
      if(n_predictors >= 2){
        vif_results <- performance::check_collinearity(best_model)
        max_vif <- max(vif_results$VIF)
        passes_vif <- all(vif_results$VIF < vif_threshold)
        
        if(!passes_vif){ # lots of text, was useful while testing.
          # cat("Excluding model with predictors:", paste(names(coef(best_model))[-1], collapse = ", "), 
          #     "- Max VIF =", round(max_vif, 2), "\n")
        }
      }
      
      if(passes_vif){
        model_results[[length(model_results) + 1]] <- list(
          model = best_model,
          aic = AIC(best_model),
          predictors = names(coef(best_model))[-1],
          formula = formula(best_model),
          max_vif = max_vif,
          n_predictors = n_predictors,
          pred_signature = paste(sort(names(coef(best_model))[-1]), collapse = "_")
        )
      }
    }
    
    # check if we have any valid models
    if(length(model_results) == 0){
      cat("No models passed VIF threshold of", vif_threshold, "\n")
    } else {
      cat("\n", length(model_results), "models passed VIF checks\n")
      
      # deduplicate models
      unique_signatures <- unique(sapply(model_results, function(x) x$pred_signature))
      unique_models <- lapply(unique_signatures, function(sig) {
        model_results[[which(sapply(model_results, function(x) x$pred_signature == sig))[1]]]
      })
      
      cat("After deduplication:", length(unique_models), "unique models\n")
      
      # extract AICs and find best
      aics <- sapply(unique_models, function(x) x$aic)
      best_aic <- min(aics)
      
      # identify models within 2 AIC units of best
      top_models_idx <- which(aics - best_aic < 2)
      top_models <- unique_models[top_models_idx]
      
      # cap at top 10 models
      max_ensemble_size <- 10
      if(length(top_models) > max_ensemble_size){
        aic_order <- order(sapply(top_models, function(x) x$aic))
        top_models <- top_models[aic_order[1:max_ensemble_size]]
        cat("Capping ensemble at", max_ensemble_size, "models\n")
      }
      
      cat("Best AIC:", best_aic, "\n")
      cat("Number of unique models within 2 AIC units:", length(top_models), "\n")
      
      # check for high variability if multiple models
      use_single_model <- FALSE
      if(length(top_models) > 1) {
        cat("\nChecking coefficient stability across ensemble candidates:\n")
        
        # thresholds for instability
        cv_threshold <- 50  # CV > 50% is concerning
        iqr_threshold <- 0.5  # IQR/mean > 0.5 is concerning
        
        all_predictors <- unique(unlist(lapply(top_models, function(x) x$predictors)))
        high_variability_found <- FALSE
        
        for(pred in all_predictors) {
          coefs <- sapply(top_models, function(x) {
            if(pred %in% names(coef(x$model))) {
              coef(x$model)[pred]
            } else {
              NA
            }
          })
          coefs <- coefs[!is.na(coefs)]
          
          if(length(coefs) > 1) {
            mean_coef <- mean(coefs)
            sd_coef <- sd(coefs)
            iqr_coef <- IQR(coefs)
            q25 <- quantile(coefs, 0.25)
            q75 <- quantile(coefs, 0.75)
            
            # calc CV only if mean is not near zero
            cv <- NA
            cv_high <- FALSE
            if(abs(mean_coef) > 0.01) {
              cv <- (sd_coef / abs(mean_coef)) * 100
              cv_high <- cv > cv_threshold
            } else {
              cv_high <- TRUE  # near-zero mean with variation is unstable
            }
            
            ## calc relative IQR
            iqr_relative <- ifelse(abs(mean_coef) > 0.01, iqr_coef / abs(mean_coef), NA)
            iqr_high <- !is.na(iqr_relative) && iqr_relative > iqr_threshold
            
            ## report stability
            cat("  ", pred, ":\n")
            if(!is.na(cv)) {
              cat("    Mean =", round(mean_coef, 3), ", SD =", round(sd_coef, 3), 
                  ", CV =", round(cv, 1), "%")
              if(cv_high) cat(" [HIGH CV!]")
              cat("\n")
            } else {
              cat("    Mean ≈ 0 (", round(mean_coef, 4), "), SD =", round(sd_coef, 3))
              cat(" [UNSTABLE - near-zero mean]\n")
            }
            
            cat("    IQR =", round(iqr_coef, 3), " (Q1 =", round(q25, 3), 
                ", Q3 =", round(q75, 3), ")")
            if(iqr_high) cat(" [HIGH IQR!]")
            cat("\n")
            
            ## flag if either metric shows high variability
            if(cv_high || iqr_high) {
              high_variability_found <- TRUE
            }
          } else if(length(coefs) == 1) {
            cat("  ", pred, ": appears in only 1 model (coef =", round(coefs, 3), ")\n")
          }
        }
        
        ## IF its unstable, then use single model 
        if(high_variability_found) {
          cat("\n⚠ High coefficient variability detected across ensemble.\n")
          cat("Selecting single best model instead of ensembling.\n")
          use_single_model <- TRUE
        } else {
          cat("\n Coefficient stability acceptable across ensemble.\n")
        }
        cat("\n")
      }
      
      ## create env_model
      if(length(top_models) == 1 || use_single_model) {
        ## single best model (either only 1 candidate or forced by instability)
        if(use_single_model) {
          ## select the single best model by AIC
          best_idx <- which.min(sapply(top_models, function(x) x$aic))
          top_models <- list(top_models[[best_idx]])
        }
        
        env_model <- top_models[[1]]$model
        env_model$scaling_params <- scaling_params
        class(env_model) <- c("scaled_glm", class(env_model))
        
        cat("Using single best model with", top_models[[1]]$n_predictors, "predictor(s):",
            paste(top_models[[1]]$predictors, collapse = ", "), "\n")
        cat("AIC =", round(top_models[[1]]$aic, 2), "\n")
        
        ## customized predict for single scaled model to be saved for l8ter
        predict.scaled_glm <- function(object, newdata = NULL, type = "response", ...) {
          if(!is.null(newdata)) {
            newdata_scaled <- newdata
            for(pred in names(object$scaling_params)) {
              if(pred %in% names(newdata)) {
                newdata_scaled[[pred]] <- (newdata[[pred]] - object$scaling_params[[pred]]$center) / 
                  object$scaling_params[[pred]]$scale
              }
            }
            predict.glm(object, newdata = newdata_scaled, type = type, ...)
          } else {
            predict.glm(object, type = type, ...)
          }
        }
      } else {
        
        ## ensemble of top models
        cat("Ensembling", length(top_models), "models\n")
        
        delta_aic <- sapply(top_models, function(x) x$aic - best_aic)
        aic_weights <- exp(-0.5 * delta_aic)
        aic_weights <- aic_weights / sum(aic_weights)
        
        for(i in seq_along(top_models)) {
          vif_str <- if(is.na(top_models[[i]]$max_vif)) "N/A" else round(top_models[[i]]$max_vif, 2)
          cat("  Model", i, "(", top_models[[i]]$n_predictors, "pred):",
              paste(top_models[[i]]$predictors, collapse = ", "),
              "(AIC =", round(top_models[[i]]$aic, 2), ", Weight =", round(aic_weights[i], 3), ")\n")
        }
        
        ## create ensemble model object
        env_model <- structure(
          list(
            models = lapply(top_models, function(x) x$model),
            weights = aic_weights,
            scaling_params = scaling_params,
            call = call("ensemble"),
            coefficients = top_models[[1]]$model$coefficients
          ),
          class = c("ensemble_glm", "list")
        )
        
        ## custom predict method for ensemble (use for later)
        predict.ensemble_glm <- function(object, newdata = NULL, type = "response", ...) {
          if(!is.null(newdata)) {
            newdata_scaled <- newdata
            for(pred in names(object$scaling_params)) {
              if(pred %in% names(newdata)) {
                newdata_scaled[[pred]] <- (newdata[[pred]] - object$scaling_params[[pred]]$center) / 
                  object$scaling_params[[pred]]$scale
              }
            }
          } else {
            newdata_scaled <- NULL
          }
          
          predictions <- sapply(object$models, function(model) {
            predict(model, newdata = newdata_scaled, type = type, ...)
          })
          
          if(is.matrix(predictions)) {
            predictions %*% object$weights
          } else {
            sum(predictions * object$weights)
          }
        }
      }
      
      # generate predictions and residuals
      phen_estimate$predicted <- predict(env_model, newdata = phen_estimate, type = "response")
      phen_estimate$residuals <- phen_estimate$onset - phen_estimate$predicted
      
      # calculate metrics
      res_env <- phen_estimate$residuals
      rmse_val <- sqrt(mean(res_env^2))
      
      # calculate R²
      ss_res <- sum(res_env^2)
      ss_tot <- sum((phen_estimate$onset - mean(phen_estimate$onset))^2)
      r2_val <- 1 - (ss_res / ss_tot)
      
      cat("R² =", round(r2_val, 3), "| RMSE =", round(rmse_val, 3), "\n")
    }
  }
  phen_estimate <- phen_estimate %>% 
    rename(predicted_onset = predicted) 
  
  
  # ensure that hopkins_delay_midpoint is available in the original raw data as well
  phen_data <- phen_data %>% 
    mutate(
      # Extract numeric values from bin intervals
      bin_lower = as.numeric(sub("\\((.+),.+\\]", "\\1", hopkins_bin)),
      bin_upper = as.numeric(sub("\\(.+,(.+)\\]", "\\1", hopkins_bin)),
      hopkins_delay_midpoint = (bin_lower + bin_upper) / 2
    )
  
  ################################################################################################
  ## Visualize Outputs 
  # Hopkins Only
  hop_plot1 <- ggplot(phen_estimate, aes(x = hopkins_delay_midpoint, y = onset)) +
    geom_point(size = 3) +
    geom_smooth(method = "lm", se = TRUE) +
    annotate("text", x = min(phen_estimate$hopkins_delay_midpoint), y = max(phen_estimate$onset),
             label = sprintf("R² = %.3f\nRMSE = %.2f days", r2_hop[[2]], rmse_val_hop),
             hjust = 0, vjust = 1) +
    labs(x = "Hopkins Expected Delay Bin (Middle Value for Bin)",
         y = "Observed Onset (day of year)",
         title = paste("Testing Hopkins Bioclimatic Law on", target_taxa[1])) +
    theme_minimal()
  hop_plot2 <- ggplot(phen_estimate, aes(x = corrected_onset_prediction, y = onset)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
    #geom_ribbon(aes(ymin = corrected_onset_lower, ymax = corrected_onset_upper), 
    #alpha = 0.2, fill = "blue") +
    geom_point(size = 3) +
    annotate("text", x = min(phen_estimate$onset), y = max(phen_estimate$corrected_onset_prediction),
             label = sprintf("R² = %.3f\nRMSE = %.2f days", r2_hop[[2]], rmse_val_hop),
             hjust = 0, vjust = 1) +
    labs(x = "Corrected Predicted Onset (anchored + actual slope)",
         y = "Empirical Estimate Onset (day of year)",
         title = paste("Testing Hopkins Bioclimatic Law on", target_taxa[1])) +
    theme_bw()
  # Env Space
  if(nrow(phen_estimate) > 3){
    env_plot <- ggplot(phen_estimate, aes(x = onset, y = predicted_onset)) +
      geom_point(alpha = 0.6) +
      geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
      annotate("text", x = min(phen_estimate$onset), y = max(phen_estimate$predicted_onset),
               label = sprintf("R² = %.3f\nRMSE = %.2f days", r2_val, rmse_val),
               hjust = 0, vjust = 1) +
      labs(x = "Observed Onset", y = "Predicted Onset", 
           title = paste("Using Climatic Vars to Predict", target_taxa[1])) +
      theme_bw()

    
  }
  
  if(nrow(phen_estimate) > 3){
    # return everything as list
    return(list(
      hopkins_model = hopkins_model, 
      env_model_onset = env_model,
      data = phen_data,
      phen_est_data = phen_estimate, 
      plots = list(hopkins = c(hop_plot1, hop_plot2), enviroment = c(env_plot)), 
      metrics = list(
        hopkins_r2 = r2_hop[[2]], 
        hopkins_rsme = rmse_val_hop, 
        env_r2_onset = r2_val, 
        env_rsme_onset = rmse_val
      )
    ))
  } else{
    # return everything as list
    return(list(
      hopkins_model = hopkins_model, 
      env_model_onset = NULL, 
      data = phen_data,
      phen_est_data = phen_estimate, 
      plots = list(hopkins = c(hop_plot1, hop_plot2), enviroment = NULL), 
      metrics = list(
        hopkins_r2 = r2_hop[[2]], 
        hopkins_rsme = rmse_val_hop, 
        env_r2 = NULL, 
        env_rsme = NULL
      )
    ))
    
  }
}

target_taxa <- "Anthophora hololeuca"

a_holo_out <- model_baseline_onset_w_clim(target_taxa, 
                                          bees_or_plants = "bees", 
                                          phenovision_flowering_data = phenovision_flowering_data,
                                          cch2_annotated_specimens = cch2_annotated_specimens, 
                                          bee_occs = bee_occs, 
                                          extra_hololeuca_data = extra_hololeuca_data, 
                                          min_doy = 21, # found by histogram
                                          max_doy = 234)

a_holo_out$plots$hopkins
a_holo_out$plots$enviroment


target_taxa <- na_taxonomy %>% 
    filter(alignedParentName == "Psorodendron arborescens") %>% 
    arrange(desc(nameStatus == "Accepted"), desc(alignedParentName == name)) %>% 
    pull(name)

psoro_a_out <- model_baseline_onset_w_clim(target_taxa, 
                                          bees_or_plants = "plants", 
                                          phenovision_flowering_data = phenovision_flowering_data,
                                          cch2_annotated_specimens = cch2_annotated_specimens, 
                                          bee_occs = bee_occs, 
                                          extra_hololeuca_data = extra_hololeuca_data, 
                                          min_doy = 30, # found by histogram
                                          max_doy = 228)

psoro_a_out$plots$hopkins
psoro_a_out$plots$enviroment

target_taxa <- na_taxonomy %>% 
    filter(alignedParentName == "Psorothamnus polydenius") %>% 
    arrange(desc(nameStatus == "Accepted"), desc(alignedParentName == name)) %>% 
    pull(name)

psoro_p_out <- model_baseline_onset_w_clim(target_taxa, 
                                          bees_or_plants = "plants", 
                                          phenovision_flowering_data = phenovision_flowering_data,
                                          cch2_annotated_specimens = cch2_annotated_specimens, 
                                          bee_occs = bee_occs, 
                                          extra_hololeuca_data = extra_hololeuca_data, 
                                          min_doy = 38, # found by histogram
                                          max_doy = 244)

psoro_p_out$plots$hopkins
psoro_p_out$plots$enviroment

target_taxa <- na_taxonomy %>% 
    filter(alignedParentName == "Psorodendron fremontii") %>% 
    arrange(desc(nameStatus == "Accepted"), desc(alignedParentName == name)) %>% 
    pull(name)

psoro_f_out <- model_baseline_onset_w_clim(target_taxa, 
                                          bees_or_plants = "plants", 
                                          phenovision_flowering_data = phenovision_flowering_data,
                                          cch2_annotated_specimens = cch2_annotated_specimens, 
                                          bee_occs = bee_occs, 
                                          extra_hololeuca_data = extra_hololeuca_data, 
                                          min_doy = 19, # found by histogram
                                          max_doy = 238)

psoro_f_out$plots$hopkins
psoro_f_out$plots$enviroment

target_taxa <- na_taxonomy %>% 
    filter(alignedParentName == "Psorothamnus emoryi") %>% 
    arrange(desc(nameStatus == "Accepted"), desc(alignedParentName == name)) %>% 
    pull(name)

psoro_e_out <- model_baseline_onset_w_clim(target_taxa, 
                                          bees_or_plants = "plants", 
                                          phenovision_flowering_data = phenovision_flowering_data,
                                          cch2_annotated_specimens = cch2_annotated_specimens, 
                                          bee_occs = bee_occs, 
                                          extra_hololeuca_data = extra_hololeuca_data, 
                                          min_doy = 12, # found by histogram
                                          max_doy = 185)

psoro_e_out$plots$hopkins
psoro_e_out$plots$enviroment

target_taxa <- na_taxonomy %>% 
    filter(alignedParentName == "Psorodendron schottii") %>% 
    arrange(desc(nameStatus == "Accepted"), desc(alignedParentName == name)) %>% 
    pull(name)

psoro_sc_out <- model_baseline_onset_w_clim(target_taxa, 
                                          bees_or_plants = "plants", 
                                          phenovision_flowering_data = phenovision_flowering_data,
                                          cch2_annotated_specimens = cch2_annotated_specimens, 
                                          bee_occs = bee_occs, 
                                          extra_hololeuca_data = extra_hololeuca_data, 
                                          min_doy = 15, # found by histogram
                                          max_doy = 215)

psoro_sc_out$plots$hopkins
psoro_sc_out$plots$enviroment

target_taxa <- na_taxonomy %>% 
    filter(alignedParentName == "Psorodendron spinosum") %>% 
    arrange(desc(nameStatus == "Accepted"), desc(alignedParentName == name)) %>% 
    pull(name)

psoro_sp_out <- model_baseline_onset_w_clim(target_taxa, 
                                          bees_or_plants = "plants", 
                                          phenovision_flowering_data = phenovision_flowering_data,
                                          cch2_annotated_specimens = cch2_annotated_specimens, 
                                          bee_occs = bee_occs, 
                                          extra_hololeuca_data = extra_hololeuca_data, 
                                          min_doy = 86, # found by histogram
                                          max_doy = 248)

psoro_sp_out$plots$hopkins
psoro_sp_out$plots$enviroment
```



